\section{Introduction}
% Motivation
Elliptic partial differential equations (PDEs) model steady-state problems such as electrical potential, heat conduction, and diffusion processes \cite{straussPartialDifferentialEquations2007}, and have applications in engineering, finance, and environmental science, among others \cite{evansPartialDifferentialEquations2010,pavliotisStochasticProcessesApplications2014}.

% Problem formulation
We study source identification constrained by an elliptic PDE: given a domain $\Omega$ with boundary $\partial\Omega$ and boundary observations $y = u|_{\partial\Omega}$ of the solution $u = \mathcal{S}f$, where $\mathcal{S}$ is the solution operator of the PDE, we seek to recover the unknown interior source $f$. This leads to an inverse problem of the form
\begin{equation}\label{eq:problem}
    \min_{f} \left\| u - y \right\|_{L^2(\partial\Omega)}^2
    + \lambda^2 \left\| W f \right\|_{L^2(\Omega)}^2
\end{equation}
subject to
\begin{eqnarray}\label{eq:pde}
    \begin{aligned}
    -\nabla \cdot \sigma\nabla u + c u &= f \quad \text{in } \Omega, \\
    \frac{\partial u}{\partial n} &= 0 \quad \text{on } \partial\Omega.
    \end{aligned}
\end{eqnarray}
Here, $\lambda$ is a regularization parameter, $W$ is the  regularization weight operator, $\sigma$ is the diffusion coefficient and $c$ is the reaction coefficient. We refer to the operator $\mathcal{K}: f \mapsto u|_{\partial\Omega}$ as the forward operator, mapping the source to the boundary values of the solution. The Neumann boundary condition in Equation~\eqref{eq:pde} ensures no flow over the boundary.

This problem arises in applications such as inverse imaging, crack determination, and electromagnetic brain mapping \cite{nielsenComputingIschemicRegions2013,elulGenesisEEG1971,PDFRecoveryCracks}. For example, in electroencephalography (EEG), it can be used for improved understanding and treatment of psychological disorders like Alzheimer’s, Parkinson's, and depression \cite{elulGenesisEEG1971, ElectromagneticBrainMapping}.

However, source identification problems of this form are ill-posed, as the operator $\mathcal K$ has a rapid decay of singular values and a large nullspace. Reconstructions are therefore unstable and non-unique \cite{hansen2010discrete}. Standard approaches like Tikhonov regularization introduce bias to the reconstructions, making solutions appear near the domain boundary regardless of the true location of the source \cite{elvetunRegularizationOperatorSource2020}. True sources are often sparse, yet Tikhonov smooths reconstructions, blurring distinct sources into connected regions.

% Prior work
Elvetun and Nielsen address the bias of Tikhonov by introducing regularization weights $W$ that use information about the forward operator $\mathcal{K}$ to correct the bias \cite{elvetunRegularizationOperatorSource2020}. This method shows promising results, correctly identifying the position of one or multiple sources from boundary data \cite{elvetunRegularizationOperatorSource2020,elvetunModifiedTikhonovRegularization2020}. However, their approach requires explicit knowledge about $\mathcal{K}$ and its SVD, making it computationally infeasible for large-scale problems. The rSVD approximates the SVD via random sampling and is computationally efficient \cite{halkoFindingStructureRandomness2011}, but still requires explicit knowledge of $\mathcal{K}$.
For sparse sources, Elvetun and Nielsen's method suffers from the same smoothing problems as Tikhonov, blurring the reconstructions and blending distinct sources \cite{elvetunModifiedTikhonovRegularization2020}. This motivates methods that can recover low-rank solutions. DLRA provides a robust framework for this \cite{schotthoferLowrankLotteryTickets2022}, but first-order methods, like Adam \cite{kingmaAdamMethodStochastic2017} and steepest-descent, converge slowly for ill-conditioned problems \cite{chong2023introduction,hansen2010discrete,nocedal2006numerical}.

% Contribution
We propose two methods that address the computational cost of Elvetun and Nielsen's weights and the convergence issues of DLRA for low-rank solutions. The first is a matrix-free variant of the rSVD that avoids explicit formation of $\mathcal{K}$. The method computes an approximate rank-$k$ SVD $K \approx U \Sigma V^T$, where $K$ is the discretization of $\mathcal{K}$, that can be used to solve the weighted Tikhonov problem for large-scale problems. It relies on applications of the forward operator $K$ and adjoint $K^*$, which can be done efficiently by pre-computing sparse Cholesky factorizations of the PDE operator. The method is applied to previously computationally infeasible problems without a meaningful loss in reconstruction quality.

Second, we apply conjugate gradient (CG) within the DLRA framework to accelerate convergence and recover sparse sources. The method adapts the rank-adaptive DLRA method proposed by Schotthöfer et al. for neural network training \cite{schotthoferLowrankLotteryTickets2022} to the setting of sparse source identification. The convex optimization problem of Equation~\eqref{eq:problem}, with Elvetun and Nielsen's $W$, defines an objective function $\Phi$ with a symmetric positive definite (SPD) Hessian $H$, for which CG is particularly well-suited \cite{hestenesMethodsConjugateGradients1952,fletcherFunctionMinimizationConjugate1964}. The resulting method converges substantially faster than DLRA with steepest descent and yields low-rank solutions that separate distinct sources more clearly than full-rank weighted Tikhonov.

The proposed methods are not limited to the particular problem studied in this thesis. The matrix-free rSVD method is well suited for a broader class of elliptic source identification problems, where the forward operators have rapid spectral decay \cite{evansPartialDifferentialEquations2010,engl1996regularization}. The DLRA-CG method can be used for low-rank problems with a quadratic objective function and an SPD Hessian.

% Thesis outline 
The remainder of this thesis is organized as follows. Section~\ref{sec:mathematical_background} covers preliminaries and describes the discretization of Equation~\eqref{eq:problem} and the weights of Elvetun and Nielsen. Section~\ref{sec:approximating} presents the matrix-free rSVD (Algorithm~\ref{alg:discrete_rsvd}) and a variant of it applied to $K^*$ (Algorithm~\ref{alg:discrete_rsvd_Kstar}). Section~\ref{sec:dynamical_low_rank_approximation} motivates DLRA and introduces the DLRA-CG method (Algorithm~\ref{alg:DLRA-CG}) and a preconditioned variant (Algorithm~\ref{alg:DLRA-PCG}). Section~\ref{sec:results} reports numerical experiments for both methods, examining computational cost, reconstruction error, and convergence rates. Section~\ref{sec:discussion} discusses the results, and Section~\ref{sec:conclusion} concludes with limitations and future directions.
