\section{Approximating the Forward Operator}
\input{figures/singular_values.tex}
% Opening: motivation and feasibility
The goal of this sections if to compute the regularization weights given in Eq.~\eqref{eq:weights_fem}. For fine meshes, the $N_b \times N$ forward matrix in Eq.~\eqref{eq:discrete_K} is not known explicitly, as storing it in memory would be infeasible. In order to compute $w_i$, one first needs one PDE solve to form $v_i = K e_i$. Forming $y_i = K^+ v_i$ requires solving the least-squares problem $\min_x \| K x - v_i \|$, which would involve further PDE solves and adjoint solves (as part of an iterative solver applying $K^T$ and $K$ to solve the normal equations). In total, it would cost $O(N)$ forward and adjoint PDE solves. For fine meshes where $N$ is large, this approach is infeasible.

This motivates the use of the rSVD. If we let $K = U \Sigma V^T$ be the SVD of $K$, we observe that $K^+ K = VV^T$, so
\begin{equation*}
    w_i = V V^T e_i.
\end{equation*}
An approximate rank-$k$ truncated SVD $K \approx U_k \Sigma_k V_k^T$ can efficiently be computed via the rSVD. The forward operator $\mathcal{K}$ is compact, since the PDE solution operator $\mathcal{S}$ smooths the source $f$, as discussed in Sec.~6.3 in in Evans, \cite{evansPartialDifferentialEquations2010}, and the singular values of $\mathcal{K}$ must therefore decay to zero \cite{engl1996regularization}. For elliptic operators, the decay of singular values is rapid, as illustrated in Fig.~\ref{fig:singular_values}

Since the forward operator $K$ has a large null-space, and rapidly decaying singular values \textcolor{red}{(why?)}, $k$ can be selected particularly small. After a rank-$k$ approximation is obtained, and the weights can be approximated as $w_i \approx V_k V_k e_i$.

\begin{itemize}
\item The computational bottleneck
\begin{itemize}
    \item Forming $K$ requires $N$ PDE solves
    \item Requires knowing $K^+$
\end{itemize}
\item  The spectral structure of K: K is compact, singular values decay rapidly due to the smoothing nature of the elliptic PDE.
\item Therefore a rank $k < N$ approximation captures the main action of K.
\item We do not know $K$ explicitly, standard rSVD needs to be modified
\item Show a singular value decay plot here.
\end{itemize}


\subsection{A matrix-free rSVD scheme}
\input{figures/rsvd.tex}
\textcolor{red}{Add oversampling parameter $p$.}
Let $K$ be the $N_b \times N$ discrete forward operator given in Eq.~\eqref{eq:K}. We are interested in a rank-$k$ approximation $K_k$ of the matrix $K$. By the Eckart–Young–Mirsky theorem, the optimal rank-$k$ approximation of $K$ is the truncated SVD obtained by keeping the $k$ first singular values, and their corresponding singular values. The randomized SVD provides an efficient alternative, by computing an approximate rank-$k$ factorization $K \approx U_k \Sigma_k V_k$, which also approximates the optimal truncated rank-$k$ SVD decomposition. We will now present a matrix-free variation of the rSVD algorithm, suited for problems where the matrix $K$ is not explicitly known.
\nomenclature[G]{$k$}{Target rank}
\nomenclature[G]{$p$}{Oversampling parameter}

Like the rSVD, this matrix-free formulation consists of two stages. In the first stage, an approximate basis $Q$ of $K$ is constructed through random sampling of the column space of $K$. In the second stage, the matrix $K$ is projected onto the subspace $\operatorname{span}(Q)$, and a SVD factorization is computed of the projected matrix.

In the first stage, we apply $K$ on $k$ random vectors $\psi_1, \psi_2, \dots, \psi_k$ drawn from some distribution, typically the normal distribution. This yields the random samples $y_1, y_2, \dots, y_k$, where $y_i = K \psi_i$. If we gather the samples and random vectors into matrices $Y = [y_i]_{i=1}^k$ and $\Psi = [\psi_i]_{i}$, we can express $Y$ as
\begin{equation*}
    Y = K \Psi.
\end{equation*}
\textcolor{red}{(something about how $y_i$ is most likely to line up with the first singular vector)}. Geometrically, we want the $i$'th sample $y_i$ to approximately point in the direction of the the $i$ right singular vector $v_i$. To remove contributions of the previous right singular vectors $v_{i-1}, v_{i-2}, \dots, v_{1}$ on $y_i$, we must orthogonalize the random samples. A QR-factorization
\begin{equation*}
    QR = Y
\end{equation*}
achieves such an orthogonalization, obtaining the matrix $Q$ which contains the orthogonalized column-vectors $q_1, q_2, \dots, q_k$. This concludes the first step, where we have obtained an approximation of the column space of $K$:
\begin{equation*}
    \operatorname{span}(Q) \approx \operatorname{span}(K)
\end{equation*}
\nomenclature[G]{$Y$}{Sample matrix}
\nomenclature[G]{$\Psi$}{Random matrix}

For the second stage, we project $K$ down to the subspace $\operatorname{span}(Q)$. This is typically done by computing $B = Q^T K$. However, since we do not know $K$ explicitly, we instead compute $B^* = K^* Q$, and then $B$ can be found by computing $B = (B^*)^*$.

To do this in practice, we use the the definition of $K^*$ under the finite element space $V_h$,
\begin{equation}\label{eq:K_star}
    (K^* z)^T M = z^T M_\partial K.
\end{equation}
If we choose $z = M_{\partial}^{-1} q_i$ and insert it into Eq.~\eqref{eq:K_star}, we see that
\begin{equation}\label{eq:rsvd_relation}
    (K^* z)^T M
    = (M_{\partial}^{-1} q_i)^T M_{\partial} K
    = q_i^T K.
\end{equation}
Hence, if we let
\begin{equation}\label{eq:rsvd_adjoint_problem}
    b_i = (K^* (M_{\partial}^{-1} q_i))^T M
\end{equation}
for $i = 1, 2, \dots, k$, and define the $k \times N$ matrix
\begin{equation*}
    B = \begin{bmatrix} b_1 \\ \vdots \\ b_k \end{bmatrix}
\end{equation*}
it must be that, by Eq.~\eqref{eq:rsvd_relation},
\begin{equation*}
    B = Q^T K,
\end{equation*}
as desired.

We have computed the projection $B = Q^T K$ by solving the adjoint problem in Eq.~\eqref{eq:rsvd_adjoint_problem} $k$ times. In the final part of the second stage, we take the SVD of $B$,
\begin{equation*}
    B = \tilde U_k \Sigma_k V_k^T,
\end{equation*}
and set $U_k = Q \tilde U_k$. This gives an approximate SVD decomposition of the form
\begin{equation}\label{eq:rsvd_approx}
    K \approx U_k \Sigma_k V_k^T.
\end{equation}
By noting that $U_k \Sigma_k V_k^T = Q \tilde U_k \Sigma_k V_k^T = Q B = Q Q^T K$, and that $\operatorname{rank}(Q) = k$, we know that the factorization in Eq.~\eqref{eq:rsvd_approx} is a rank-$k$ approximation of $K$. The procedure is summarized in Algorithm~\ref{alg:discrete_rsvd}.

\input{algorithms/matrix-free-rsvd.tex}


\subsection{Approximation of the regularization weights}
Let $K_k = U_k \Sigma_k V_k^T$ be the rank-$k$ approximate SVD of $K$ obtained using Algorithm~\ref{alg:discrete_rsvd}. The Tikhonov weights $W = \operatorname{diag}(w_1, \dots, N)$, given in Equation~\eqref{eq:weights_fem}, can then be approximated by setting
\begin{equation}\label{eq:rsvd_weights}
    y_i = V_k V_k^T e_i.
\end{equation}
To see this, let $K = U \Sigma V^T$ be the SVD of $K$ and we recall that $K^+ = V \Sigma^+ U^T$ (see Appendix~\textcolor{red}{(ref)}). Since $U_k$ is orthogonal, and since $\Sigma_k^+ \Sigma_k = I_k$ (\textcolor{red}{(need details)}), we have that $y = K_k^+ K_k e_i$ simplifies to Equation~\eqref{eq:rsvd_weights}.
