\section{Dynamical low-rank approximation}
\input{figures/low_rank_manifold.tex}
Consider the problem of finding a rank-$r$ solution $X$ via the update scheme $X^{(i+1)} = X^{(i)} + \alpha_i D^{(i)}$, where $D^{(i)}$ is a search direction and $\alpha_i > 0$ a step size. Starting from a rank-$r$ matrix $X^{(0)}$, there is in general no guarantee that $D^{(i)}$ preserves low rank, so that $X^{(i)}$ may leave the space $\mathcal{M}_r$ of rank-$r$ matrices after a single step.

One can replace $D^{(i)}$ by its projection onto the tangent space $\mathcal{T}_{X^{(i)}}(\mathcal{M}_r)$. This is the idea behind the Dynamical Low-Rank Approximation (DLRA) of Koch and Lubich \cite{kochDynamicalLowrankApproximation2007}. They proposed, in a continuous-time setting, evolving an approximation on $\mathcal{M}_r$ by projecting the derivative of the matrix onto the tangent space at the current approximation. Using this concept in our discrete setting, we replace $D^{(i)}$ by its tangent-space projection $\mathcal{P}_{X^{(i)}}[D^{(i)}]$, yielding the update
\begin{equation*}
    X^{(i+1)} = X^{(i)} 
    + \alpha_i \mathcal{P}_{X^{(i)}}(D^{(i)}).
\end{equation*}
This is illustrated in Figure~\ref{fig:low_rank_manifold}. Rather than updating $X^{(i)}$ directly, the DLRA algorithm operates on its SVD factorization $X^{(i)} = U_x \Sigma_x V_x^T$, updating the factors $U_x$, $\Sigma_x$, and $V_x$ at each iteration (we use the subscript ``$x$'' to
distinguish these factors from those of the forward operator
$K = U \Sigma V^T$).

However, as pointed out by Koch and Lubich, when $\Sigma_x$ is ill-conditioned, which occurs when $r$ exceeds the effective rank of $X^{(i)}$, the projector $\mathcal{P}_{X^{(i)}}$ suffers from stiffness, leading to very small step sizes and issues with rounding errors.

To address this, Lubich and Oseledets proposed splitting the projector $\mathcal{P}_{X^{(i)}}$ into simpler sub-projections \cite{lubichProjectorsplittingIntegratorDynamical2013}. Crucially, each sub-projection can be evaluated without forming $\Sigma^{-1}$, and avoids the stiffness problem present in the original formulation. In recent years, additional advancements have been made to DLRA, offering improved stability, parallelism, and rank-adaptivity \cite{cerutiUnconventionalRobustIntegrator2020, cerutiRankadaptiveRobustIntegrator2021}.

Historically, DLRA was developed for time-continuous matrix differential equations. Schotthöfer et al. recently proposed a discrete rank-adaptive update scheme based on a robust projector-splitting integrator for training low-rank neural networks \cite{schotthoferLowrankLotteryTickets2022}, see Algorithm~\ref{alg:dlra}. The discrete DLRA scheme used in this work follows their formulation.
\input{algorithms/DLRA.tex}


\subsection{Step-size selection and rank adaptation}
Algorithm~\ref{alg:dlra} admits several choices for the step size $\alpha_i$ and the truncation criterion.

\paragraph{Step size.}
A fixed step size $\alpha_i = \alpha > 0$ coincides with one step of gradient descent on the factors $K'$, $L'$, and
$\hat \Sigma$, analogous to the stochastic gradient descent (SGD) step used by Schotthöfer et al. \cite{schotthoferLowrankLotteryTickets2022}, but using the full gradient. Adam computes adaptive step sizes from estimates of the first and second moments of the gradient \cite{kingmaAdamMethodStochastic2017}, and Schotthöfer et al. find it converges faster in practice than SGD for training neural networks \cite{schotthoferLowrankLotteryTickets2022}. For the quadratic objective $\Phi$ considered in this thesis (Equation~\eqref{eq:discrete_inverse_problem}), steepest descent selects the step size that minimizes $\Phi$ along $D^{(i)}$,
\begin{equation*}
    \alpha_i
    = \frac{\|\nabla\Phi(X^{(i)})\|_F^2}
           {\langle D^{(i)},\, H D^{(i)} \rangle_F},
\end{equation*}
and requires no tuning (Section~8.2 in Chong et al. \cite{chong2023introduction}).


\paragraph{Rank adaption.} In the dynamical low-rank update (KLS) step of Algorithm~\ref{alg:dlra}, the factors $\hat{U}$ and $\hat{V}$ have $2r$ columns and $\hat{\Sigma}$ is of shape $2r \times 2r$. The truncation step (line 10) moves the factors back to a specified maximum rank $r$ \cite{schotthoferLowrankLotteryTickets2022}. The truncation can also be done adaptively via a truncation threshold $\vartheta$, as done by Schotthöfer et al \cite{schotthoferLowrankLotteryTickets2022}. 


\subsection{Application to the inverse problem}\label{sec:application_to_inverse_problem}
In many applications, the source $f \in L^2(\Omega)$ in Equation~\eqref{eq:problem} is sparse \cite{elvetunWeightedSparsityRegularization2023}, i.e., of low rank. If $x \in \mathbb{R}^{n^2}$ is some discretization of $f$, reordering $x$ into an $n \times n$ matrix $X = \operatorname{mat}(x)$, we expect that $r = \operatorname{rank}(X)$ is a small number ($r \ll n$). It is therefore natural to consider low-rank methods which can locate such solutions.

In the following sections, we consider the problem of minimizing a quadratic objective function
\begin{equation*}
    \Phi(X): \mathbb{R}^{n \times n} \rightarrow \mathbb R.
\end{equation*}
In our discrete case, the minimization problem given in Equation~\eqref{eq:problem} translates into minimizing
\begin{equation*}
    \Phi(X) = \frac 1 2 \|Kx - y\|_{M_\partial}^2
    + \frac{\lambda^2}{2} \| W x \|_M^2.
\end{equation*}
Using the product rule, we get the gradient (with respect to $x$) $\nabla \Phi (X) = K^T M_\partial (K x - y) + \lambda^2 W^T M W x$, and the Hessian $H = K^T M_\partial K + \lambda^2 W^T M W$. The terms $K^T M_\partial K$ and $\lambda^2 W^T M W$ are SPD, making the Hessian $H$ SPD too.


\subsection{DLRA with conjugate search directions}
Since $\Phi$ is quadratic with a SPD Hessian $H$, the conjugate gradient (CG) method is a natural minimization strategy \cite{saadIterativeMethodsSparse2003,fletcherFunctionMinimizationConjugate1964,hestenesMethodsConjugateGradients1952}.

CG generates a sequence of search directions $\{D^{(i)}\}$ that are $H$-conjugate, i.e.\ $\langle D^{(i)}, H D^{(j)}\rangle_F = 0$ for $i \neq j$. Starting from the initial search direction $D^{(0)} = -G^{(0)}$, each step computes an optimal step size
\begin{equation*}
    \alpha_k = \frac{\|G^{(i)}\|_F^2}{\langle D^{(i)},\, H D^{(i)}\rangle_F},
\end{equation*}
and updates the search direction by the Fletcher-Reeves rule \cite{fletcherFunctionMinimizationConjugate1964},
\begin{equation}\label{eq:fletcher-reeves}
    D^{(i+1)} = -G^{(i+1)} + \beta_i D^{(i)},
    \qquad
    \beta_i = \frac{\|G^{(i+1)}\|_F^2}{\|G^{(i)}\|_F^2}.
\end{equation}

In our low-rank setting, we propose replacing the descent directions used in the method of Schotthöfer et al. (Algorithm~\ref{alg:DLRA-CG}) with $H$-conjugate search directions $\{D^{(i)}\}$ obtained via the CG method. The motivation is that $H$-conjugate directions exploit curvature information of $H$, leading to faster convergence. The full scheme is given in Algorithm~\ref{alg:DLRA-CG}, which we refer to as DLRA-CG throughout.
\input{algorithms/DLRA-CG.tex}


\subsubsection{Restarted CG}\label{sec:restarted_cg}
During the construction of $H$-conjugate search directions $\{D^{(i)}\}$, numerical round-off errors accumulate, so that 
\begin{equation*}
    \langle D^{(i)},\, HD^{(j)} \rangle_F = 0, \quad i \neq j,
\end{equation*}
will not be exactly satisfied \cite{hestenesMethodsConjugateGradients1952}. As $|i - j|$ increases, the violation of $H$-conjugacy may grow, so that the final iterate $X^{(i)}$ is not a satisfactory approximation of the solution \cite{hestenesMethodsConjugateGradients1952}. For the DLRA-CG scheme, the loss of $H$-conjugacy is accelerated by the truncation step: the update in Equation~\eqref{eq:fletcher-reeves} is derived under the assumption that $X^{(i+1)} = X^{(i)} + \alpha_i D^{(i)}$, which is not generally true after truncation.

Restarted CG \cite{powellRestartProceduresConjugate1977} addresses this by re-computing the true gradient
$G^{(i)} = \nabla\Phi(X^{(i)})$ and resetting the search direction to $D^{(i)} = -G^{(i)}$ every $N_r$ iterations, discarding the previous search directions that have lost $H$-conjugacy. In Algorithm~\ref{alg:DLRA-CG}, this replaces the recurrence $G^{(i+1)} = G^{(i)} + \alpha_i HD^{(i)}$ with a fresh gradient evaluation $G^{(i+1)} = \nabla\Phi(X^{(i+1)})$, and sets $D^{(i+1)} = -G^{(i+1)}$, skipping the $\beta_i$ step.


\subsection{DLRA using preconditioned CG}
Preconditioning is a standard technique for accelerating CG methods, referred to as preconditioned CG (PCG). The system $H x = y$ is replaced by the equivalent system $P^{-1} H x = P^{-1} y$, where $P \approx H$ is the preconditioner matrix. The convergence of CG depends on $\kappa(H)$ (see Sec.~6.11.3 in Saad) \cite{saadIterativeMethodsSparse2003}, so $P$ is chosen such that $\kappa(P^{-1} H) \ll \kappa(H)$. For elliptic PDE-based forward operators $K$, the Hessian $H = K^T M_\partial K + \lambda^2 W^T M W$ typically has a large condition number, making preconditioning effective.

In the DLRA-CG scheme, the preconditioned system $P^{-1} H x = P^{-1} y$ leads to the following adjustments in Algorithm \ref{alg:DLRA-CG}:
\begin{equation*}
    \alpha_i = \frac
    {\langle G^{(i)}, Z^{(i)} \rangle_F}
    {\langle D^{(i)}, H D^{(i)} \rangle_F}, \quad
    \beta_i = \frac
    {\langle G^{(i+1)}, Z^{(i+1)} \rangle_F}
    {\langle G^{(i)}, Z^{(i)} \rangle_F}, \quad
    D^{(i+1)} = -Z^{(i+1)} + \beta_i D^{(i)}.
\end{equation*}
These modifications are equivalent to standard preconditioned conjugate gradient, as originally described by Concus et al. \cite{concusGeneralizedConjugateGradient1976}. The resulting scheme summarized in Algorithm \ref{alg:DLRA-PCG}, which we refer to as DLRA-PCG.
\nomenclature[H]{$\Phi$}{Objective function}
\nomenclature[H]{$H$}{Hessian of $\Phi$}
\nomenclature[H]{$P$}{Preconditioner matrix}
\input{algorithms/DLRA-PCG.tex}


\subsubsection{Choice of preconditioner}
The effectiveness of PCG depends on the choice of $P$. A good preconditioner must approximate $H$ well, but must be cheap to apply at each iteration. We consider three common preconditioners: Jacobi scaling, incomplete Cholesky (IC), and (sparse) Cholesky. Since $H =  K^T M_\partial K + \lambda^2 W^T M W$ has a sparse plus low-rank structure, the Cholesky-based preconditioners are modified using the Woodbury formula to exploit this \cite{woodbury1950inverting}. We refer to Golub and Van Loan for a comprehensive treatment of preconditioning and iterative methods \cite{golubMatrixComputations4th2013}.

\paragraph{Jacobi scaling.} A Jacobi preconditioner $P = \operatorname{diag}(H)$ approximates $H$ by its diagonal. The approximation is cheap to compute, as $P^{-1}$ is the inverse of a diagonal matrix. It is most effective when $H$ is diagonally dominant.
 
\paragraph{Incomplete Cholesky.} For a symmetric positive definite (SPD) matrix $S$, the incomplete Cholesky factorization computes an approximation $S \approx  L L^T$, where $L$ is lower triangular. The Hessian $H = K^T M_\partial K + \lambda^2 W^T M W$ consists of a low-rank part plus a sparse part. We only apply the IC factorization to the sparse part $S = \lambda^2 W^T M W$, yielding $S \approx L L^T$. To account for the low-rank part, we let $K = U \Sigma V^T$ be the SVD of $K$ (which is assumed to be of rank $k$), and define $F = V \Sigma L_U$, where $L_U$ is the Cholesky factor of the $k \times k$ matrix $U^T M_\partial U$ (which is SPD since $M_\partial$ is SPD and $U$ has linearly independent columns). Then, the low-rank part can be written as $F F^T = V \Sigma L_U L_U^T \Sigma V^T = K^T M_\partial K$. By the Woodbury formula,
\begin{equation*}
    (S + FF^T)^{-1} = S^{-1} - 
    S^{-1} F (I_k + F^T S^{-1} F)^{-1} F^T S^{-1}.
\end{equation*}
Replacing $S^{-1}$ with the approximate inverse $L^{-T} L^{-1}$ yields the preconditioner
\begin{equation*}
    P^{-1} = L^{-T} L^{-1} -
    L^{-T} (L^{-1} F) C^{-1} (L^{-1} F)^T L^{-1},
\end{equation*}
where $C = I_k + (L^{-1} F)^T(L^{-1} F)$ is a $k \times k$ SPD matrix.

\paragraph{Sparse Cholesky.} The sparse Cholesky preconditioner follows the same Woodbury construction as the incomplete Cholesky preconditioner, but replaces the approximate factorization of $S$ with the exact sparse Cholesky factorization $S = LL^T$. Since the Woodbury formula then gives $P^{-1} = H^{-1}$, this is equivalent to solving $H x = y$ directly \textcolor{red}{(wrong)}, and CG converges in a single iteration. However, the setup is more expensive compared to incomplete Cholesky.
