\section{Results}
\subsection{Experimental setup}\label{sec:problems}
\input{figures/problem_illustration.tex}
Figure~\ref{fig:problem_illustration} shows the three meshes and source functions used throughout the results. Problem I is a square mesh with a single source, used as a baseline. Problem II uses the same geometry with three sources to assess reconstruction under multiple sources. Problem III replaces the square domain with an ``L''-shaped mesh and two sources, introducing a non-convex geometry.

For all experiments, we set the diffusion coefficient $\sigma = 1$ and the reaction coefficient $c = 1$ in Equation~\eqref{eq:pde}. The finite element space $V_h$ (Equation~\eqref{eq:fem_space}) uses a basis of continuous piecewise-linear hat functions $\phi_i$ (continuous Galerkin, degree 1).

Unless stated otherwise, we use $N = 64^2 = 4096$ for Problems I and II and $N = 6127$ for Problem III, regularization parameter $\lambda = 10^{-4}$, oversampling $p = 5$, and Gaussian random test vectors $\psi_i$ for the rSVD.


\subsection{Computational setup}
All numerical experiments were performed on a laptop running Fedora Linux 43 with Linux kernel version 6.19. The system was running with an Intel(R) Core(TM) i7-10510U (4 cores, 8 threads) with 16GB of memory.

Python 3.9 was used for all computations, together with NumPy and SciPy for linear algebra routines \cite{harrisArrayProgrammingNumPy2020,virtanenSciPy10Fundamental2020}. The FEniCS software was used for solving PDEs using the finite element method \cite{alnaesFEniCSProjectVersion2015,loggAutomatedSolutionDifferential2012}. All library specifications, algorithm implementations, and notebooks used to generate the results in this thesis are openly available at: \href{https://github.com/EliasSev/masters-thesis}{github.com/EliasSev/masters-thesis}.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Matrix-free Randomized SVD Part 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Matrix-free rSVD}
\subsubsection{Computational time}\label{sec:computational_time}
\input{figures/rsvd_computational_time_comp.tex}
\input{tables/rsvd_computational_time_comp.tex}
The left plot of Figure~\ref{fig:rsvd_computational_time_comp} shows the computational time of Algorithm~\ref{alg:discrete_rsvd} for target ranks $k = 10,\, 25$ and $50$, compared to forming $K = T A^{-1} M$ explicitly (Section~\ref{sec:approximating}) and taking its SVD. Table~\ref{tab:rsvd_computational_time_comp} shows the relative speedup comparing different methods, with notation $t^{(k)}_{\text{rSVD}}$ for the time of using Algorithm~\ref{alg:discrete_rsvd} at target rank $k$.

The time $t_K$ surpasses $t_{\text{rSVD}}$ for $N < 1000$ across the selected target ranks, and $t_{\text{SVD}}$ for $N < 5000$. The time $t_{K}$ increases more rapidly than $t_{\text{rSVD}}$ for all target ranks, and $t_{\text{SVD}}$ is observed to increase faster than $t_{K}$, matching their respective theoretical costs of $\mathcal{O}(N_b^2 N)$ and $\mathcal{O}(N_b N \log N$), established in Section~\ref{sec:approximating}.
 
The computational times $t_{\text{rSVD}}$ for target ranks $10$, $25$ and $50$ are seen to scale near linearly on this range of $N$ (right plot of Figure~\ref{fig:rsvd_computational_time_comp}). The rate of increase depends on the target rank, with a higher target rank resulting in a more rapid increase. For $N>10{,}000$, increasing $k$ from 10 to 50 results in an increase of $t_{\text{rSVD}}$ by a factor of approximately 1.9, and increasing $k$ from 10 to 25 has a factor of approximately 1.3. 


\subsubsection{Computational breakdown}\label{sec:computational_breakdown}
\input{figures/rsvd_computational_breakdown.tex}
\input{tables/rsvd_computational_breakdown.tex}
Figure~\ref{fig:rsvd_computational_breakdown} shows the median computational time for each stage of Algorithm~\ref{alg:discrete_rsvd} for selected $N$ and increasing $k$, with the relative computational cost shown in the bottom row. The median is taken over 30 runs for $N \le 40{,}000$ and 10 runs for $N = 400{,}000$. The `Setup' stage includes the assembly and factorization of the mass matrices $M$ and $M_\partial$, and the setup of the $K$ and $K^*$ solvers (Section~\ref{sec:initial_matrix_factorizations}). The remaining stages of Algorithm~\ref{alg:discrete_rsvd} are `2. $Y = K\Psi$' (lines 1--6), `3. $QR = Y$' (lines 7--8), `4. $B = Q^T K$' (lines 9--14), and `5. SVD' (lines 15--18). Table~\ref{tab:rsvd_computational_breakdown} shows relative computational time for selected target ranks and mesh sizes.

The dominant costs of Algorithm~\ref{alg:discrete_rsvd} are the forward solves using $K$ (stage 2) and the adjoint solves using $K^*$ (stage 4), both of which scale roughly linearly with $N$, matching the theoretical cost of $\mathcal{O}(l N \log N)$ with $l = k + p$ (Section~\ref{sec:initial_matrix_factorizations}). The setup stage has theoretical cost $\mathcal{O}(N^{3/2})$ but appears near-linear in $N$ over the range considered here. Its share of total time decreases with $k$, because setup is independent of the target rank.

Stage 5's relative cost is primarily determined by the target rank $k$, but also scales linearly with $N$, matching the predicted $\mathcal{O}(l^2 N)$ cost of the SVD of $B$. Its share grows with $k$: at $N = 400{,}000$, stage 5 accounts for 3\% of total time at $k = 10$ and 36\% at $k = 200$. Stage 3 (the QR factorization of the $N_b \times l$ matrix $Y$) remains below 1\% of total time across all mesh sizes.


\subsubsection{Target rank effect on reconstruction quality}\label{sec:target_rank_reconstruction}
\input{figures/test_problems_selected.tex}
Figure~\ref{fig:test_problems_selected} shows the weighted Tikhonov solution of Equation~\eqref{eq:discrete_inverse_problem}: the left panel uses the exact $K$ and $W$, while the middle and right panels use $K$ and $W$ approximated by Algorithm~\ref{alg:discrete_rsvd} at target ranks $k = 10$ and $k = 50$.

For Problem~I (Figure~\ref{fig:test_problem_I}) at $k = 10$, the location of the source is partially identified (middle), slightly shifted towards the interior and with more smoothing than the exact Tikhonov solution (left). The weighted Tikhonov regularization itself already produces significant smoothing \cite{elvetunRegularizationOperatorSource2020}, and low-rank approximations amplifies this effect at small $k$. At $k = 50$ (right), the reconstruction is visually identical to the exact Tikhonov solution. In Problems~II and~III (Figures~\ref{fig:test_problem_II} and~\ref{fig:test_problem_III}), $k = 10$ shows the same over-smoothing and shift of each source. These issues disappear at $k = 50$. Across all three problems, $k = 50$ matches the exact Tikhonov reconstruction visually at a small fraction of the computational cost of solving the exact problem (Section~\ref{sec:computational_time}).

\input{figures/rsvd_over_k.tex}
\input{tables/rsvd_over_k.tex}
Figure~\ref{fig:rsvd_over_k} shows the reconstruction error for target ranks $k \in [5,\, 100]$ and oversampling parameters $p = 0,\, 5,\, 10$, measured by the Euclidean norm, Earth Mover's distance (EMD), and area under the curve of intersection over union (AUC-IoU). See Appendix~\ref{appendix:performance_metrics} for details on each metric. An AUC-IoU score of 1 indicates a perfect overlap of the reconstruction and the true source. The error curves flatten beyond $k \approx 100$, so the figure is cropped there for more detail at smaller $k$. Table~\ref{tab:rsvd-over-k} gives the relative difference (in \%) between the rSVD and exact Tikhonov solutions at selected $k$ (up to 199), for $p = 5$, across all three problems and metrics.

The rSVD reconstruction error approaches the exact reconstruction error as $k$ increases, for all three error metrics. Larger oversampling $p$ yields a small but consistent reduction in error across all three metrics, most visible in the EMD curves. At $k = 50$ and $p = 5$, the Euclidean relative difference is 0.56\%, 2.44\%, and 2.92\% for Problems~I, II, and III; the AUC-IoU gives 1.97\%, 2.87\%, and 4.07\%; and the EMD gives 4.74\%, 6.44\%, and 39.13\%. The EMD is more sensitive to low-amplitude noise in the reconstruction, which increases its relative error even though the noise is difficult to see in Figure~\ref{fig:test_problems_selected}. At $k = 199$ (the largest rank considered), Euclidean and AUC-IoU are within fractions of a percent of the exact Tikhonov values. The EMD ranges from 0.12\% to 7.85\%, which indicates its sensitivity to small noise or small differences shape or position compared to the exact Tikhonov solution.


\subsubsection{Test vector distribution effect on error}
\input{figures/random_test_vectors.tex}
Three test vector distributions are compared: standard Gaussian ($\psi_{ij} \sim \mathcal{N}(0,1)$), Rademacher ($\psi_{ij} \in \{-\tfrac{1}{2}, \tfrac{1}{2}\}$ with equal probability, see \cite{montgomery-smithDistributionRademacherSums1990}), and a spatially localized Gaussian peak
\begin{equation*}
    \psi_i(x) = \exp\left(-\frac{|x - x_0|^2}{2\sigma^2}\right),
\end{equation*}
where the center of the peak $x_0$ is sampled uniformly from the domain and $\sigma = 0.1$. Examples are shown in Figure~\ref{fig:random_test_vectors}.

\input{figures/random_vectors_error.tex}
\input{figures/clipped_random_vectors.tex}
Figure~\ref{fig:random_vectors_error} compares reconstruction error across the three distributions for Problem~I (similar results hold for Problems~II and III), using the Euclidean norm, EMD, and
AUC-IoU. Each curve shows the mean over 50 runs with $\pm 1$ standard deviation. Rademacher and standard Gaussian yield nearly identical errors across all three metrics. The Gaussian peak produces
lower EMD error and smaller run-to-run variance. Figure~\ref{fig:clipped_random_vectors} shows individual reconstructions with the colormap clipped to $0.05$, revealing that the peak distribution
generates less low-amplitude noise than Gaussian or Rademacher.


\subsubsection{Mesh resolution effect on reconstruction error}
\input{figures/mesh_resolution_effect.tex}
Figure~\ref{fig:mesh_resolution_effect} shows the relative Euclidean, EMD, and AUC-IoU errors for rSVD (Algorithm~\ref{alg:discrete_rsvd}) at fixed target ranks $k = 10,\, 25,\, 50$ as the mesh resolution $N$ increases, averaged over 5 repetitions

For fixed $k$, the Euclidean error and AUC-IoU remain constant across mesh resolutions; reconstruction quality does not depend on $N$. A higher target rank gives lower error throughout. The EMD shows a slight decrease with $N$, likely because finer meshes produce smoother spatial transitions between neighboring mesh elements.

Although $\operatorname{rank}(K) = N_b$ grows with $N$, the target rank $k$ need not be adjusted: refining the mesh adds no new information about the source $f$.


\subsubsection{Approximation of singular values}
\input{figures/approximation_of_singular_values.tex}
\input{tables/singular_values_comparison.tex}
Figure~\ref{fig:approximation_of_singular_values} shows $\sigma_i / \sigma_i^{\text{(true)}}$ for the 100 largest singular values of $K$ (Problem~I), with $\sigma_i$ computed by rSVD applied to $K$ or $K^*$ (Algorithms~\ref{alg:discrete_rsvd} and~\ref{alg:discrete_rsvd_Kstar}) for $p \in \{0, 5, 10\}$. Each curve is the mean over 20 runs, with $\pm 1$ standard deviation. Table~\ref{tab:singular_values_comparison} lists selected values.

The singular values are under-estimated across all oversampling parameters, with the bias increasing for
higher-index singular values. A larger $p$ reduces the underestimation. Applying rSVD on $K^*$, the quality of the singular values worsens the approximation quality.

As was found in Section~\ref{sec:target_rank_reconstruction}, a target rank of $k = 50$ yields reconstruction errors within 1\% to 4\% (Euclidean and AUC-IoU) of the true Tikhonov solution, and visually indistinguishable reconstructions. At this rank with $p = 5$, the ratio $\sigma_{50}/\sigma_{50}^{(\text{true})}$ is $0.969$ for rSVD on $K$ and $0.929$ for rSVD on $K^*$.


\subsubsection{Computational time of adjoint operator}
\input{tables/computational_time_adjoint.tex}
Table~\ref{tab:computational_time_adjoint} shows that rSVD on $K^*$ takes approximately 1.3 times longer than rSVD on $K$ for $k = 10$, $1.2$ times longer for $k = 50$, and $1.1$ times longer for $k = 200$, across all reported $N$. When the recovery of the SVD is included ($t_2 / t$), it takes approximately 1.3 times as long across all $k$ and $N$.

For larger $k$, the recovery of the SVD takes up an increasing amount of time, as it involves QR factorizations of an $N_b \times k$ and an $N \times k$ matrix, in addition to an SVD of a $k \times k$ matrix (Section~\ref{sec:reconvering_the_svd}). Additionally, as $k$ goes to $200$, the ratio $t_1 / t$ gets closer to 1.0, as an increasing amount of the time of Algorithm~\ref{alg:discrete_rsvd} goes to the SVD step (Section~\ref{sec:computational_breakdown}).


\subsubsection{Reconstruction error using the adjoint operator}
\input{figures/rsvd_adjoint_error.tex}
\input{tables/rsvd_adjoint_error.tex}
\input{figures/clipped_solutions.tex}
Figure~\ref{fig:rsvd_adjoint_error} shows the reconstruction error of the weighted Tikhonov solution where $K$ and $W$ are approximated using rSVD on $K$ (Algorithm~\ref{alg:discrete_rsvd}) and rSVD on $K^*$ (Algorithm~\ref{alg:discrete_rsvd_Kstar}) for target ranks $k \in [10, 100)$. The error is measured in Euclidean norm, EMD, and AUC-IoU. Table~\ref{tab:rsvd_adjoint_error} reports the relative difference of rSVD on $K$ compared to rSVD on $K^*$, and Figure~\ref{fig:clipped_solutions} shows solutions of both methods at $k=15$ with the colormap clipped to expose low-amplitude noise in the reconstructions.

In terms of Euclidean norm, both methods give practically the same error, within $\pm 1$\% across all problems and target ranks. For AUC-IoU, the relative difference at $k = 10$ is between $-6$\% and $-4$\%, and and between -2\% and 1\% for $k \ge 25$ with no consistent direction.

The EMD shows that at $k = 10$, rSVD on $K$ gives 10\% to 19\% worse reconstructions than rSVD on $K^*$, though at this rank the smoothing and positional shift of the reconstruction are already significant (Section~\ref{sec:target_rank_reconstruction}). In the range $k \approx 20$ to $k \approx 50$, rSVD on $K^*$ performs worse: at $k=25$, Algorithm~\ref{alg:discrete_rsvd} performs 9\% to 31\% better across all three problems. The difference is a result of noise: at $k=25$, reconstructions from rSVD on $K^*$ contain more low-amplitude noise than those from rSVD on $K$ (Figures~\ref{fig:clipped_solution_K_star} and~\ref{fig:clipped_solution_K}), to which the EMD is particularly sensitive.


\subsubsection{Accuracy on noisy data}
\input{figures/noisy_problem.tex}
\input{tables/noisy_problem.tex}
\input{figures/reconstruction_noisy_problem.tex}
We measure the reconstruction error for the transformed weighted Tikhonov problem of Equation~\eqref{eq:transformed_discrete_inverse_problem}. We replace $y = K x$ with $\tilde y = y + \epsilon$, where $\epsilon_i \sim \mathcal{N}(0, 1)$ and $\tfrac{\| \epsilon \|}{\| y \|} = \delta$ with $\delta = 0.01$. The regularization parameter is set to $\lambda = \lambda_{\text{dp}}$, found using the discrepancy principle with safety factor $\nu_{\text{dp}} = 10$ (see Section~5.2 in Hansen \cite{hansen2010discrete}), so that $\lambda$ is selected independently for each method and rank.

Figure~\ref{fig:noisy_problem} shows reconstruction errors when $K$ and $W$ are approximated using rSVD (Algorithm~\ref{alg:discrete_rsvd}) and tSVD (Section~\ref{sec:tsvd}) at increasing ranks $k$. The error is measured in Euclidean norm, EMD and AUC-IoU. The optimal rank $k^*$ is marked with vertical lines, and listed in Table~\ref{tab:noisy}. Figure~\ref{fig:reconstruction_noisy_problem} shows reconstructions for Problem~II at selected ranks, illustrating how a target rank that is too high includes noise in the reconstruction.

In terms of Euclidean error, the optimal rank is between 50 and 85 for rSVD, and 45 and 85 for tSVD. The EMD is more sensitive to higher ranks which include more noise (Figure~\ref{fig:reconstruction_noisy_problem}), and the optimal rank for EMD is between 20 and 30 for both methods. For AUC-IoU, optimal ranks are between 50 and 100 for rSVD, and between 65 and 95 for tSVD, except for Problem~III with rSVD where the optimal rank exceeds the tested range ($k \leq 100$).

At their respective optimal ranks, rSVD and tSVD give nearly identical Euclidean errors and AUC-IoU scores across the three problems, with differences below 0.25\%. The largest gap is in EMD: tSVD achieves 2.35\% to 6.33\% lower EMD, with the largest difference on Problem~III.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dynamical Low-Rank Approximation using (Preconditioned) Conjugate Gradient Part 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dynamical low-rank approximation solvers}
Unless stated otherwise, all experiments use a rank-$50$ rSVD approximation of $K$ and $W$ (Algorithm~\ref{alg:discrete_rsvd}, $p = 5$, standard Gaussian test vectors $\psi_i$) to form $\Phi$, its gradient, and its Hessian (Section~\ref{sec:application_to_inverse_problem}), on which Algorithm~\ref{alg:DLRA-CG} is applied to find a low-rank solution.

We write $X^* \in \mathbb{R}^{n \times n}$ for the matrix form of the true source, i.e., $x^* = \operatorname{vec}(X^*)$, and $N_r$ for the restart period of restarted DLRA-CG (Section~\ref{sec:restarted_cg}).


\subsubsection{Rate of convergence}
\input{figures/number_of_iterations.tex}
\input{tables/convergence_iterations.tex}
Figure~\ref{fig:number_of_iterations} compares the reconstruction error per iteration for three DLRA step-size rules (Algorithm~\ref{alg:dlra}) and DLRA-CG (Algorithm~\ref{alg:DLRA-CG}), with and without restarting. The right panel shows the effect of the restart period $N_r$ on the convergence of DLRA-CG.

DLRA-CG reaches the lowest error fastest but then rises, a consequence of the loss of $H$-conjugacy described in Section~\ref{sec:restarted_cg}, illustrating a form of semi-convergence. Restarted DLRA-CG (with $N_r = 10$) avoids this rise, at the cost of slower convergence, but convergences faster than the three DLRA step-size rules, which all fail to converge within $25{,}000$ iterations (and the Adam rule gets stuck).

A smaller $N_r$ slows convergence but drives the error lower, and for larger $N_r$, the semi-convergent behavior of DLRA-CG re-appears. Table~\ref{tab:convergence_iterations} shows the iteration count to reach selected error thresholds. DLRA-CG reaches a relative error of $0.9$ in 49 iterations, and restarted DLRA-CG takes 36 iterations. The step-size rules Adam, fixed, and steepest descent reaches this threshold in 1757, 21532, and 1532 iterations. To reach a threshold of $0.75$, it takes DLRA-CG without and with restarting 100 and 160 iterations respectively, the three other methods fail to reach this error within the iteration budget of $25{,}000$. Only restarted DLRA-CG reaches the relative error threshold of $0.70$.


\subsubsection{Quality of the low-rank solutions}
\input{figures/low_rank_convergence.tex}
\input{figures/low_rank_restarted_cg.tex}
DLRA-CG shows semi-convergence: the iterates initially approach the true source, but start to drift away from the optimal solution when $H$-conjugacy is lost (Section~\ref{sec:restarted_cg}). Figure~\ref{fig:low_rank_convergence} shows the difference between the minimum reconstruction iterate and the converged iterate. In both Problems~I and II, the positions of the sources are slightly shifted away from the boundary for the converged solutions (right panels), while the optimal iterate is correctly located (middle panels). In both Problems~I and II, the late-stage iterate (right panels) shows a slight inward shift of the source positions, while the minimum-error iterate (middle panels) correctly reproduces the source locations. The early iterate (left panels) has not yet localized the sources.

Compared to the full-rank solutions of Section~\ref{sec:target_rank_reconstruction}, which have smoothing pulled toward the interior, the low-rank solutions are more uniformly smoothed with no such tail. For Problem~I, this creates a circular source, and for Problem~II, this results in cleanly separated sources (in Figure~\ref{fig:test_problem_II} they were connected). Additionally, at $k = 50$, the peak amplitudes are $0.28$ (rank-1 DLRA-CG) versus $0.26$ (full-rank rSVD) for ProblemI, and $0.64$ versus $0.48$ for Problem~II, where the upper-left source is weaker in the rank-2 solution. The true source amplitude is 1.0 (Figure~\ref{fig:problem_illustration}). The low-amplitude noise in the full-rank rSVD solution is replaced by a checker-like pattern of similar amplitude.

Restarted DLRA-CG ($N_r = 10$) achieves a lower relative error than DLRA-CG without restarting (Figure~\ref{fig:number_of_iterations}, left). The converged solutions in Figure~\ref{fig:low_rank_restarted_cg} are correctly located and less diffuse, with peak amplitudes of $0.40$ and $0.80$ for Problems~I and II, closer to the true amplitude of $1.0$ than both the full-rank rSVD Tikhonov solutions and the solutions without restarting.


\subsubsection{Effect of rank on reconstructions}
\input{figures/low_rank_different_r.tex}
\input{figures/low_rank_solution_error.tex}
Figure~\ref{fig:low_rank_different_r} shows DLRA-CG solutions for Problems~I and II at ranks $r = 1, 2, 5$. For Problem~I, the reconstruction error increases with rank, and at rank 5 the solution resembles the full-rank Tikhonov reconstruction (Figure~\ref{fig:test_problem_I}). For Problem~II, rank 2 is optimal; at rank 5 the reconstruction similarly resembles the full-rank case (Figure~\ref{fig:test_problem_II}).

Figure~\ref{fig:low_rank_solution_error} shows that for Problem~I, the rank 1 DLRA-CG solution achieves lower Euclidean and AUC-IoU error than the full rank rSVD solution. For EMD, the full-rank solution is better, possibly because the focused distribution of the rank-1 solution differs more from the flat true source than the smoother full-rank reconstruction. For ProblemII, the rank-2 solution outperforms rank-1, rank-5, and full-rank solutions across all metrics for $k > 10$. Reconstructions of rank 5 achieves approximately the same error as full-rank rSVD reconstructions.

As shown in Section~\ref{sec:target_rank_reconstruction}, full-rank rSVD solutions approach but do not outperform the exact Tikhonov solution in the range $k \in [5,\,100]$ (Figure~\ref{fig:rsvd_over_k_II}). The low-rank solutions do: for Problem~I, the rank-1 solution achieves lower Euclidean and AUC-IoU error than the exact Tikhonov solution, and for Problem~II, the rank-2 solution outperforms it across all metrics \textcolor{red}{[TODO: at which $k$ does this happen?]}.


\subsubsection{Effect of different preconditioners}
\input{figures/precondition_times.tex}
