\section{Appendix}


\subsection{Functional analysis and the FEM}\label{appendix:functional_analysis_and_the_fem}
\paragraph{Lebesgue space $L^2$}
Let $\Omega \subset \mathbb{R}^d$ be an open set and $\mathbb{F}$ denote either $\mathbb{R}$ or $\mathbb{C}$. The space $L^2(\Omega)$, often referred to as the Lebesgue space of square integrable functions, is defined as 
\begin{equation*}
    L^2(\Omega) =
    \left\{
        f : \Omega \to \mathbb{F} \, \Bigg| \,
        \int_{\Omega} |f(x)|^2 \, dx < \infty
    \right\},
\end{equation*}
with inner product
\begin{equation*}
    \langle f,\, g \rangle_{L^2(\Omega)} = \int_\Omega f(x) \overline{g(x)} \, dx.
\end{equation*}
The space $L^2(\Omega)$ is a Hilbert space, i.e., a complete inner product space. For details, see Section~3 in Kreyszig \cite{kreyszigIntroductoryFunctionalAnalysis1989}.


\paragraph{Sobolev space $H^1$}
The space $H^1(\Omega)$, known as the first-order Sobolev space, is defined as
\begin{equation*}
    H^1(\Omega) =
    \left\{
        f \in L^2(\Omega) \, \Bigg| \,
        \frac{\partial f}{\partial x_i} \in L^2(\Omega), \, i = 1, \dots, d  
    \right\},
\end{equation*}
with inner product
\begin{equation*}
    \langle f,\, g \rangle_{H^1(\Omega)} = 
    \langle f,\, g \rangle_{L^2(\Omega)} +
    \langle \nabla f,\, \nabla g \rangle_{L^2(\Omega)}.
\end{equation*}
The space $H^1(\Omega)$ is a Hilbert space and a subspace of $L^2(\Omega)$. See Section~2 in Quarteroni for further details \cite{quarteroniNumericalModelsDifferential2014}.


\paragraph{Weak derivative}
Let $\Omega \subset \mathbb{R}^d$ be an open set and $u \in L^2(\Omega)$. A function $v$ is said to be the weak derivative of $u$ if
\begin{equation*}
    \int_\Omega u(x) \frac{\partial \phi}{\partial x_i} \, dx =
    - \int_\Omega v \phi \, dx
\end{equation*}
for all infinitely differentiable $\phi$ with compact support on $\Omega$. For details, see Section~2 in Quarteroni \cite{quarteroniNumericalModelsDifferential2014}.


\paragraph{The stiffness and the mass matrix}\label{appendix:stiffness}
Let $V_h \subset H^1(\Omega)$ be the finite element space defined in Equation~\eqref{eq:fem_space}. Under the basis $\{\phi_i\}_{i=1}^N$, the stiffness matrix $A \in \mathbb{R}^{N \times N}$ and mass matrix $M \in \mathbb{R}^{N \times N}$ are defined as
\begin{equation*}
    A_{ij} = \int_\Omega 
    \left( \sigma \nabla \phi_j \cdot \nabla \phi_i 
    + c \phi_j \phi_i \right) dx,
    \quad
    M_{ij} = \int_\Omega \phi_j \phi_i \, dx.
\end{equation*}
Both matrices are symmetric positive definite (Appendix~\ref{sec:positive_definiteness}) under the standing assumptions that $\sigma$ is positive definite and $c > 0$, see Theorem~3.4 and Theorem~4.10 in Larson and Bengzon \cite{larsonFiniteElementMethod2013}. The boundary mass matrix $M_\partial \in \mathbb{R}^{N_b \times N_b}$ is given by
\begin{equation*}
    (M_\partial)_{ij} = \int_{\partial\Omega} \phi_i \phi_j \, dx,
\end{equation*}
where the indices $i, j$ tabulate over the boundary indices of the mesh. Like the mass matrix $M$, the boundary mass matrix $M_\partial$ is SPD \cite{larsonFiniteElementMethod2013}.

$A$, $M$ and $M_\partial$ are sparse matrices, since $A_{ij}$, $M_{ij}$ and $(M_\partial)_{ij}$ are non-zero only if $\phi_i$ and $\phi_j$ share an element in the mesh.

The assembly of stiffness matrices and mass matrices in two dimensions can be done through methods of numerical integration, as described in Section~3.5 in Larson and Bengzon \cite{larsonFiniteElementMethod2013}.


\subsection{Linear algebra and matrix decompositions}

\paragraph{Symmetric positive definiteness}\label{sec:positive_definiteness}
Let $A \in \mathbb{R}^{n \times n}$. $A$ is said to be symmetric if $A = A^T$. If $A$ is symmetric and if 
\begin{equation*}
    x^T A x > 0
\end{equation*}
for all $x \in \mathbb{R}^n$, then $A$ is said to be symmetric positive definite (SPD). For details, see Section~4.2 in \cite{golubMatrixComputations4th2013}.


\paragraph{QR factorization}
The QR factorization of $A\in\mathbb R^{m\times n}$ is a decomposition
\begin{equation*}
    A = QR,
\end{equation*}
where $Q\in\mathbb R^{m\times n}$ has orthonormal columns and $R\in\mathbb R^{n\times n}$ is upper triangular. The matrix $Q$ forms an orthonormal basis for $\operatorname{range}(A)$, as described in Section 6 of Golub and Van Loan \cite{golubMatrixComputations4th2013}.


\paragraph{LU decomposition}
The LU decomposition of $A \in \mathbb{R}^{n \times n}$ is a factorization
\begin{equation*}
    A = L U,
\end{equation*}
where $L$ is a lower triangular matrix and $U$ an upper triangular matrix. $A$ has an LU decomposition under the condition that $\operatorname{det}(A[1:k, 1:k]) \neq 0$ for $k=1,\dots, n-1$ (see Theorem~3.2.1 in Golub and Van Loan \cite{golubMatrixComputations4th2013}). More generally, any square matrix $A$ has a LU decomposition with partial pivoting (LUP), $PA = LU$. Computing the LU decomposition costs $\mathcal{O}(n^3)$ flops, specifically, $\tfrac{2}{3}n^3$.

Given the LU decomposition $A = LU$, the system $Ax = y$ can be solved by solving $Lz = y$ and $Ux = z$ with $z = Ux$. The two triangular systems are easily solved in $\mathcal{O}(n^2)$ flops.


\paragraph{Cholesky factorization}
Assume that $A\in\mathbb R^{n\times n}$ is SPD. Then $A$ has a unique Cholesky factorization
\begin{equation*}
    A = L L^T,
\end{equation*}
where $L$ is a lower triangular matrix with positive diagonal elements. This is a special case of the LU decomposition, ith $U = L^T$. The cost of this procedure is $\mathcal{O}(n^3)$, specifically, $\tfrac{1}{3}n^3$ flops, making it faster than standard LU decomposition. For details, see Section~4.2 in Golub and Van Loan \cite{golubMatrixComputations4th2013}.


\paragraph{Incomplete Cholesky factorization}
If $A \in \mathbb{R}^{n\times n}$ is SPD with Cholesky factorization $A = L L^T$, then the incomplete Cholesky factorization is an approximation
\begin{equation*}
    A \approx \tilde L \tilde L^T,
\end{equation*}
where $\tilde L$ is a lower triangular matrix which is sparser than $L$. There are many strategies for computing the incomplete Cholesky, see Lin and Moré \cite{linIncompleteCholeskyFactorizations1999}. The main advantage over the full Cholesky factorization is reduced storage and computational cost, making it particularly effective for large sparse SPD matrices.


\subsection{Solving triangular systems}\label{appendix:triangular}
Let $T \in \mathbb{R}^{n \times n}$ be triangular. Solving $T x = y$ is done by forward (or backward) substitution: starting from the first (or last) row, each $x_i$ is solved directly after substituting all previously computed values. If $T$ is sparse, zero entries are skipped, and by Lemma~2.2.1 in George and Liu \cite{georgeComputerSolutionLarge1984} the cost reduces to $\mathcal{O}(\operatorname{nnz}(T))$, where $\operatorname{nnz}(T)$ is the number of nonzeros in $T$.


\paragraph{General linear systems}
Let $A \in \mathbb{R}^{n\times n}$ be invertible and consider $A x = y$, then the solution is $x = A^{-1} y$. In practice, the standard procedure is to compute the LU factorization $A = LU$, reducing the problem to two triangular systems
\begin{equation*}
    L z = y, \quad U x = z \implies x = U^{-1} z = U^{-1} L^{-1} y.
\end{equation*}
Each triangular system can be solved by forward and backward substitution at $\mathcal{O}(n^2)$. The factorization itself costs $\tfrac{2}{3}n^3$ flops, so the problem scales as $\mathcal{O}(n^3)$. If $A$ is symmetric positive definite, $U = L^T$ and the Cholesky factorization $A = LL^T$ halves the cost to $\tfrac{1}{3}n^3$.


\paragraph{Sparse linear systems}
For sparse systems $A x = y$, computing the $LU$ factorization comes with additional challenges. Matrix entries which are zero in $A$ but non-zero in $L$ or $U$, so called fill-ins, require more memory and result in a more expensive factorization \cite{larsonFiniteElementMethod2013}. Additionally, the cost of solving the resulting triangular systems is reduced. Much research has gone into the development algorithms which avoid fill-ins, for instance, see Davis \cite{davisDirectMethodsSparse2006}.


\paragraph{The nested dissection method}
For sparse SPD matrices arising from 2D elliptic PDEs (mass matrices, stiffness matrix), the structure of the matrices can be exploited. Alan George proposed the nested dissection method in 1973 \cite{georgeNestedDissectionRegular1973}, a divide and conquer approach, that computes Cholesky factorization in $\mathcal{O}(N^{3/2})$ flops. The resulting triangular matrix $L$ contains $\mathcal{O}(N \log N)$ non-zero elements, so solving $L x = y$ costs $\mathcal{O}(\operatorname{nnz}(L)) = \mathcal{O}(N \log N)$.


\subsection{Performance metrics}\label{appendix:performance_metrics}
Let $x \in \mathbb{R}^N$ denote the true source (it's coefficient vector) and $\hat x \in \mathbb{R}^N$ it's corresponding reconstruction.

To evaluate reconstruction quality, we consider three performance metrics: Euclidean error, Earth mover’s distance (EMD), and the integrated (area under curve) intersection over union (AUC-IoU). These metrics emphasize the localization and spatial shape of the reconstructed source, rather than just the amplitude.


\paragraph{Earth mover’s distance}
The earth mover's distance measures the dissimilarity between two distributions, which in our case refers to $x$ and $\hat x$. Let $x_d$ and $\hat x_d$ be the normalized distributions of $x$ and $\hat x$, respectively, computed as
\begin{equation*}
    (x_d)_i = \frac{|x_i|}{\sum_{j=1}^N |x_j|}, \quad
    (\hat x_d)_i = \frac{|\hat x_i|}{\sum_{j=1}^N |\hat x_j|}.
\end{equation*}
The EDM between them is
\begin{equation*}
    \operatorname{EDM} =
    \inf_{\gamma_{ij} \in \Pi}
    \gamma_{ij} \sum_{ij}^N d(p_i, p_j)
\end{equation*}
where $\Pi$ is the set of all joint distributions with marginals $x_d$ and $\hat x_d$ \cite{rubnerMetricDistributionsApplications1998}.


\paragraph{AUC-IoU}
The IoU, known as the Jaccard index in statistics, measures the similarity between two sets. It is defined as the size of intersection divided by the size of the union \cite{raschkaPythonMachineLearning}.

We define the binary mask of $x$ at threshold $\tau \in (0, 1)$ as the set
\begin{equation*}
    S_{\tau}(x) = \{ i : x_i \ge \tau \operatorname{max}(x) \}.
\end{equation*}
Then, the IoU of $x$ and $\hat x$ at threshold $\tau$ is then the IoU of their binary masks,
\begin{equation*}
    \operatorname{IoU}(\tau) =
    \frac{|S_\tau(x) \cap S_\tau(\hat x)|}{|S_\tau(x) \cup S_\tau(\hat x)|}.
\end{equation*}
Here, $| \cdot |$ denotes the number of elements of a set (its cardinality).

The AUC-IoU is then the integration of the $\operatorname{IoU}(\tau)$ curve,
\begin{equation*}
    \operatorname{AUC-IoU} = \int_{0}^{1} \operatorname{IoU}(\tau) \, d\tau.
\end{equation*}
This can be approximated using the Trapezoidal rule. The AUC-IoU lies in $[0, 1]$, with 1 being a perfect score.


\subsection{Derivation of the adjoint formulation}\label{appendix:adjoint_derivation}
Consider the forward operator $\mathcal{K} = \mathcal{S} \circ \mathcal{T}$ given by Equation~\eqref{eq:K}, and the variational formulation of $\mathcal{S}$ by Equation~\eqref{eq:S_cont}. Define the bilinear form
\begin{equation*}
    a(u,\, v) =
    \int_\Omega (\sigma \nabla u \cdot \nabla v + c u v) \, dx,
\end{equation*}
then, the variational formulation of $\mathcal{S}$ reads as follows: find $u \in H^1(\Omega)$ such that
\begin{equation}\label{eq:bilinear_form}
    a(u,\, v) = \int_\Omega f v \, dx.
\end{equation}

For some $g \in L^2(\partial\Omega)$, we seek $w = \mathcal{K}^* g$. By the definition of $K^*$,
\begin{equation*}
    \langle \mathcal{K} f,\, g \rangle_{L^2(\partial\Omega)} =
    \langle f,\, \mathcal{K}^* g \rangle_{L^2(\Omega)}.
\end{equation*}
and since $\mathcal{K} f = u|_{\partial\Omega}$, we have that
\begin{equation}\label{eq:adjoint_identity_1}
    \langle \mathcal{K} f,\, g \rangle_{L^2(\partial\Omega)} =
    \int_{\partial\Omega} u g \, ds.
\end{equation}
In Equation~\eqref{eq:bilinear_form}, choose $v = w = \mathcal{K}^* g$. We see that
\begin{equation}\label{eq:adjoint_identity_2}
    a(u, w) = \int_{\Omega} f w \, dx = \langle f,\, \mathcal{K}^* g \rangle_{L^2(\Omega)}.
\end{equation}
Insert Equation~\eqref{eq:adjoint_identity_1} and \eqref{eq:adjoint_identity_2} into the definition of $\mathcal{K}^*$, and using the identity $a^*(v,\, u) = a(u,\, v)$ (by the definition of $a^*$ \cite{kreyszigIntroductoryFunctionalAnalysis1989}),
\begin{equation}\label{eq:adjoint_identity_3}
    a^*(w,\, u) = \int_{\partial\Omega} g u \, ds.
\end{equation}
Since Equation~\eqref{eq:adjoint_identity_3} must hold for all $u \in H^1(\Omega)$, it defines a variational problem for $w$. And since $a$ is symmetric ($\sigma$ is symmetric),
$a^*(w,\, u) = a(u,\, w) = a(w,\, u)$, we get that Equation~\eqref{eq:adjoint_identity_3} reduces to:
find $w \in H^1(\Omega)$ such that
\begin{equation*}
    a(w,\, u) = \int_{\partial\Omega} g u \, ds
    \quad \text{for all } u \in H^1(\Omega).
\end{equation*}
where
\begin{equation*}
    a(w, u) = \int_{\partial\Omega}
    (\sigma \nabla w \cdot \nabla u + c w u)\, dx.
\end{equation*}
This is the variational formulation of Equation~\eqref{eq:variational_formulation_adjoint} with $u := v$.
