\documentclass[a4paper,masters,en,listoffigures,listoftables]{NMBU}

% If you'd like to use chapters instead of sections, add 'book' to the options, as \documentclass[book,...]{NMBU}. Recall that you now have to use \chapter{} and then \section{}, \subsection{}, ... 

%\usepackage[nohyperlinks]{acronym} % If you need a List of Acronyms, consider using this package

% \usepackage[font=small]{caption} % if you prefer smaller captions
% \captionsetup{width=0.9\linewidth} 

% Word of advice: When you are done with your thesis, change the compiler in Overleaf to XeLaTeX! This will make your fonts and title page look much nicer than in pdfLaTeX. Just go to "Menu", in the "Settings", find "Compiler" and change to XeLaTeX.

% The NMBU style package is designed to help you write the cover and back page automatically. Make sure to include the information below to generate your title. You can set the language you are writing in with "en" for english, "bm" for bokmål, or "nn" for nynorsk.

% A table of contents is automatically included. If you have more than one figure, you need a list of figures by adding "listoffigures" to the documentclass parameters. If you have more than one table, you need a list of tables by adding "listoftables" to the documentclass parameters.

% Additional packages
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[intoc,refpage]{nomencl}

\makenomenclature
\renewcommand{\nomname}{List of Symbols}
\renewcommand{\pagedeclaration}[1]{\dotfill #1}
\setlength{\nomitemsep}{0pt}
\newcommand{\nomgroupname}[1]{%
  \ifstrequal{#1}{A}{Geometry}{}%
  \ifstrequal{#1}{B}{PDE coefficients}{}%
  \ifstrequal{#1}{C}{Continuous operators}{}%
  \ifstrequal{#1}{D}{Function spaces}{}%
  \ifstrequal{#1}{E}{Discrete operators \& matrices}{}%
  \ifstrequal{#1}{F}{Regularization parameters}{}%
  \ifstrequal{#1}{G}{Randomized SVD}{}
}
\renewcommand{\nomgroup}[1]{%
  \item[\bfseries \nomgroupname{#1}]
}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\renewcommand{\v}[1]{\boldsymbol{#1}}

% If you would like to have a bibliography with named authors, organized alphabetically, add "namedauthors" to the parameters of the document class.
\addbibresource{bib.bib} % here you include the bibliography file. 
\graphicspath{{../figures/}}

% If you are using Norwegian dates in your bibliography, add "norwegiandates" to the document class.

% When you finish your thesis, switch the LaTeX compiler from pdfLaTeX to XelaTeX (on Overleaf, "Menu"->"Compiler"). This is for 2 reasons: 1) to produce your front page with the Arial font; 2) XeLaTeX produces better fonts and (usually) smaller PDFs. Every kilobyte of space we save helps reduce CO2 emissions :)

% Good luck with your thesis! If you get in trouble with LaTeX stuff, write to me: leonardo.rydin@gmail.com

% In your .tex file, you need to include the following information to produce the first page of your thesis

\title{Matrix-Free Randomized Sketching for Column-Norm Estimation of Discretized PDE Operator}
\author{Elias Jegervatn Severinsen}
\thesisyear{2026}
\credits{30}
\faculty{Faculty of Science and Technology}
\studyprogramme{Data Science}
\supervisor{Ole Løseth Elvetun, Jonas Kusch} % Include the name of the supervisor or supervisors (this is added to the metadata of the PDF)
%\engtitle{} % Only if you write a thesis in a Scandinavian language

% You should also use the "\abstract" and "\sammendrag" to produce those. If you are writing in a Scandinavian language, the sammendrag will appear first.
\abstract{This thesis uses randomized linear algebra for estimation of a discrete PDE operator.}
\sammendrag{Denne masteroppgaven bruker randomisert lineær algebra for mastrise-fri estimering av PDE-operatorer.}
%\acknowledgements{} You can add acknowledgements here. In Scandinavian languages, use "\foreword{}" (or if in English you prefer Foreword to Acknowledgements). The foreword comes before the abstract.

\begin{document}

% If you need a List of Acronyms, place them before the \startthesis command. You will need "\usepackage[nohyperlinks]{acronym}", see line 4.

%\section*{List of Symbols}
% \begin{acronym}
%   \acro{TSO}{Transmission System Operator}
%   \acro{LSTM}{Long Short-Term Memory}
%\end{acronym}

\printnomenclature

\startthesis % This command will separate the Roman-numbered pages above and the Arabic-numbered pages below as well as create the table of contents, list of figures, list of tables, and potentially list of acronyms

\section{Introduction}
Partial differential equations (PDEs) are essential in modern modelling. They have applications in finance, physics, medicine, and biology, to name a few. The discretization of PDEs, which enables numerical solutions of such equations, can produce linear systems of equations with a large number of unknowns. For instance, in modern applications, it is not uncommon with billions of unknowns. It is therefore essential with efficient algorithms to handle large scale linear systems of equations.

In this thesis, we study the class of elliptic PDEs, which typically occur in time-independent problems. This class of PDEs show up in steady-state problems, stochastic processes, and the diffusion and flow a quantity in a region of space \cite{pavliotisStochasticProcessesApplications2014, evansPartialDifferentialEquations2010}.

In particular, we study the inverse problem of the form
\begin{equation}\label{eq:problem}
    \min_{f} \left\| \mathcal{K} f - y \right\|_{L^2(\partial\Omega)}^2
    + \lambda^2 \left\| W f \right\|_{L^2(\Omega)}^2
\end{equation}
subject to
\begin{eqnarray}\label{eq:pde}
    \begin{aligned}
    -\nabla \cdot \sigma\nabla u + c u &= f \quad \text{in } \Omega, \\
    \frac{\partial u}{\partial \v n} &= 0 \quad \text{on } \partial\Omega.
    \end{aligned}
\end{eqnarray}
Here, $\mathcal{K}$ is the forward operator which maps the source $f$ to the boundary values $\mathcal{K}f = u|_{\partial\Omega}$, $\sigma$ is the diffusion coefficient, $c$ is the reaction coefficient, and $y$ is the observed boundary data. This problem has numerous applications, such as inverse imaging, crack determination, and electromagnetic brain mapping \cite{elvetun_regularization_2020}. This problem is a source identification task: we are given observed data $y$, and seek to find the original source term $f$ such that $\mathcal K f = y$.

However, source identification problems of this form are ill-posed, as the operator $\mathcal K$ has a large null space. Consequently, solutions are not uniquely determined by the data. Moreover, small perturbations the observed data may result in large variations in the reconstructed source. Due to these challenges, typical techniques such as Tikhonov regularization fail to recover a meaningful source term.

Due to the structure of the forward operator $\mathcal K$, Tikhonov reconstructions tend to appear on or near the boundary of the domain, corresponding to directions which appear near the null space of the operator. To address this issue, Elvetun and Nielsen proposed a weighted formulation which uses knowledge about the operator \cite{elvetun_regularization_2020}. The resulting weights $W$ are constructed in a way to punish solutions which are located near the boundary. This method shows promising results, with reconstructions being located correctly, but with smoothing artifacts and a loss of the original source magnitude.

However, the construction of the weights $W$ is computationally expensive. Each column of $W$ is obtained by solving \eqref{eq:pde} with a corresponding basis function as right-hand side. Thus, the method requires one PDE solve per mesh node (or degree of freedom) in the discretization of $\Omega$, hence, the computation cost grows with the resolution of the mesh. In two spatial dimensions, the number of mesh nodes, denoted $N$, increases quadratically with the mesh resolution, whereas in three dimensions it increases cubically. This makes the method computationally infeasible for fine discretizations or large domains.

This address this issue, we propose a low-rank approximation strategy based on the randomized singular value decomposition (rSVD). The key idea is to use random sampling of $\mathcal{K}$ to find an approximate rank-$p$ SVD $K_p = U \Sigma V^T$, which then can be used to effectively construct the weights $W$. Due to the large null space of $\mathcal{K}$, and it's rapid decay of singular values, the target rank $p$ can be selected much smaller than $N$.

% In addition, I use a dynamic low-rank solver to find low-rank solutions (rank 1), and explore different additional regularization techniques (TV regularization, penalty methods)

% However, the methods presented in this thesis are not limited to this particular equation, and may be applied to elliptic PDEs in general.]


\clearpage
\section{Mathematical background}

\subsection{The forward operator}
\input{figures/forward_operator.tex}

Let $\Omega \subset \mathbb R^2$ be a bounded domain with boundary $\partial \Omega$, and consider the boundary value problem (BVP) in Equation~\eqref{eq:pde}. The coefficient $\sigma = \sigma(x) > 0$ denotes the diffusion coefficient and determines the rate and direction of diffusion in the problem. It may be either scalar-valued or matrix-valued. In the latter case, we assume that $\sigma(x)$ is symmetric and positive definite (see Appendix~\ref{sec:positive_definiteness}) to keep the problem elliptic. The reaction coefficient $c = c(x) > 0$ describes the absorption or decay within the problem. Finally, $f = f(x)$ denotes the source term, which is the unknown input of the inverse problem.
\nomenclature[A]{$\Omega$}{Computational domain}
\nomenclature[A]{$\partial\Omega$}{Boundary of the domain}
\nomenclature[B]{$\sigma,\, \sigma(x)$}{Diffusion coefficient}
\nomenclature[B]{$c,\, c(x)$}{Reaction coefficient}

Equation~\eqref{eq:pde} is a linear diffusion-reaction problem with Neumann boundary conditions, and defines a mapping
\begin{equation}
    \mathcal S: L^2(\Omega) \rightarrow H^1(\Omega), \quad \mathcal{S} f = u.
\end{equation}
Here, $L^2(\Omega)$ is the space of square integrable functions and $H^1(\Omega) \subset L^2(\Omega)$ the first-order Sobolev space, see Appendix~\ref{appendix:function_spaces}
\nomenclature[C]{$\mathcal{S}$}{PDE operator}
\nomenclature[D]{$L^2(\Omega)$}{Space of square integrable functions}
\nomenclature[D]{$H^1(\Omega)$}{First-order Sobolev space}

To model boundary observations, we introduce the trace operator
\begin{equation}\label{eq:trace}
    \mathcal T: H^1(\Omega) \rightarrow L^2(\partial\Omega), \quad \mathcal{T} u = u|_{\partial\Omega}
\end{equation}
\nomenclature[C]{$\mathcal{T}$}{Trace operator}
The term $\mathcal T u$ represents the observed values of $u$ on the boundary of $\Omega$.

The forward operator is then defined as the composition
\begin{equation}\label{eq:K}
    \mathcal K = \mathcal T \circ \mathcal S, \quad \mathcal{K} f = u|_{\partial\Omega},
\end{equation}
\nomenclature[C]{$\mathcal{K}$}{Forward operator}
which is a mapping from the source function $f$ to the observed output $u|_{\partial\Omega} = \mathcal K f$ on the boundary $\partial \Omega$. See Figure~\ref{fig:forward_operator}.

The associated inverse problem can now be formulated as follows: given boundary data $u|_{\partial\Omega}$, determine a source function $f$ such that
\begin{equation}
    \mathcal{K} f = u|_{\partial\Omega}.
\end{equation}
In order to solve such a problem computationally, the problem must be discretized. In this thesis, the discretization is done using the finite element method (FEM), described in the following sections.


\subsection{The finite element method}
\input{figures/triangulation.tex}
Consider the PDE in Eq.~\eqref{eq:pde}. Multiplying the equation by a test function $v \in H^1(\Omega)$, integrating over $\Omega$, and applying Green's identity yields the weak formulation: find $u \in H^1(\Omega)$ such that
\begin{equation}\label{eq:variational_formulation}
    \int_{\Omega} (\sigma \nabla u \cdot \nabla v + k u v) \, dx = \int_{\Omega} f v \space \, dx,
    \text{ for all } v \in H^1(\Omega).
\end{equation}
For details on the derivation, see Section~4 in Larson and Bengzon \cite{larson_finite_2013}.

Let $\mathcal M$ be a triangulation of $\Omega$ made up of $N$ nodes, with $N_b$ boundary nodes, see Fig.~\ref{fig:triangulation} as an example. Introduce the finite element space $V_h \subset H^1(\Omega)$,
\begin{equation}\label{eq:fem_space}
    V_h = \{v \in C^0(\Omega) : v|_M \text{ is linear for all } M \in \mathcal M \}.
\end{equation}
Let $\{\phi_1, \dots, \phi_N\}$ denote a basis for $V_h$, so that
\begin{equation}
    V_h = \operatorname{span}\{ \phi_1, \dots, \phi_N \}.
\end{equation}
The basis functions $\phi_i$ are piecewise linear hat functions associated with the nodes of the mesh (see \cite[p.~51]{larson_finite_2013}).

\nomenclature[A]{$\mathcal{M}$}{Triangulation (mesh)}
\nomenclature[A]{$N$}{Number of mesh nodes}
\nomenclature[A]{$N_b$}{Number of boundary mesh nodes}
\nomenclature[D]{$V_h$}{Finite element space}

The finite element approximation of \eqref{eq:variational_formulation} then reads: 
find $u_h \in V_h$ such that
 \begin{equation}\label{eq:fem_formulation}
    \int_{\Omega} (\sigma \nabla u_h \cdot \nabla v + k u_h v) \, dx = \int_{\Omega} f v \space \, dx,
    \text{ for all } v \in V_h.
\end{equation}


\subsection{The discrete forward operator}
To obtain a discrete representation of the forward operator 
$\mathcal{K} = \mathcal{T} \circ \mathcal{S}$, 
we discretize both the PDE operator $\mathcal{S}$ and the trace operator $\mathcal{T}$ using the finite element space $V_h$.

Let $f_h, u|_{\partial\Omega, h} \in V_h$ denote the finite element approximations of $f$ and $u|_{\partial\Omega}$, respectively. Since $V_h = \operatorname{span}\{\phi_1,\dots,\phi_N\}$, they can be written as
\begin{equation}\label{eq:uh_and_fh}
    u_h = \sum_{j=1}^{N} u_j \phi_j, 
    \quad
    f_h = \sum_{j=1}^{N} x_j \phi_j,
\end{equation}
with coefficient vectors $\v u = (u_1,\dots,u_N)^T$ 
and $\v x = (x_1,\dots,x_N)^T$.

Inserting \eqref{eq:uh_and_fh} into the finite element formulation \eqref{eq:fem_formulation} and testing with the basis functions 
$\{\phi_i\}_{i=1}^N$ yields the linear system
\begin{equation}\label{eq:matrix_system}
    A \mathbf{u} = M \mathbf{x},
\end{equation}
where $A \in \mathbb{R}^{N \times N}$ is the stiffness matrix and $M \in \mathbb{R}^{N \times N}$ the mass matrix, see Appendix \textcolor{red}{(ref)} for details. Equation~\eqref{eq:matrix_system} implies the discrete PDE operator $S$ takes the form
\begin{equation}
    S = A^{-1} M \in \mathbb{R}^{N \times N}
\end{equation}
Note that $S$ is a mapping of coefficients vectors: mapping $\v x$ to $\v u$.

The trace operator is discretized by a matrix 
$T \in \mathbb{R}^{N_b \times N}$ that extracts the boundary coefficients. If $\mathbf{y}$ denotes the vector of boundary coefficients, then
\begin{equation}
    \v y = T \v u.
\end{equation}

Combining the discrete PDE and trace operators yields the matrix representation of the forward operator,
\begin{equation}\label{eq:discrete_K}
    K = T S = T A^{-1} M \in \mathbb{R}^{N_b \times N}.
\end{equation}

\nomenclature[E]{$M_{dx}$}{Volume mass matrix}
\nomenclature[E]{$M_{ds}$}{Boundary mass matrix}
\nomenclature[E]{$K$}{Discrete forward operator}

\subsection{Discrete inner product and norm}
Let $V_h$ be the finite element space introduced in Eq.~\eqref{eq:fem_space} and let $f_h, g_h \in V_h$, with coefficient vectors $\v x, \v y \in \mathbb{R}^N$, respectively. The $L^2(\Omega)$ inner product (see Eq.~\eqref{eq:L2_inner_product}) of $f_h$ and $g_h$, can be written as
\begin{equation}
    \langle f_h, g_h \rangle_{L^2(\Omega)} = \v x^T M \v y,
\end{equation}
where $M$ is the mass matrix. 

This induces a discrete inner product on $\mathbb{R}^N$, with associated norm, defined by
\begin{equation}
    \langle \v x, \v y \rangle_M = \v x^T M \v y,
    \quad
    \| \v x \|_M = \sqrt{\v x^T M \v x}.
\end{equation}

Similarly, let $\v u, \v v \in \mathbb{R}^{N_b}$ 
denote coefficient vectors corresponding to the traces 
$u_h|_{\partial\Omega}$ and $v_h|_{\partial\Omega}$. The $L^2(\partial\Omega)$ inner product induces a discrete inner product and norm on $\mathbb{R}^{N_b}$, given by
\begin{equation}
    \langle \mathbf{u}, \mathbf{v} \rangle_{M_{\partial}}
    = \mathbf{u}^T M_{\partial} \mathbf{v},
    \quad
    \| \v  u \|_{M_{\partial}}
    =
    \sqrt{ \v  u^T M_{\partial} \v u},
\end{equation}
where $M_\partial$ denotes the boundary mass matrix, see Appendix~\textcolor{red}{(reference)}.

%In practice, this system is often **ill-posed** due to the large null space of $K$ and the accumulation of numerical errors in $A^{-1}$. As a result, naive solutions of Eq.~\eqref{eq:inverse_problem} are highly sensitive to noise in $\v y$, motivating the need for regularization techniques. In the following section, we discuss classical Tikhonov regularization and introduce the weighted approach that will be explored in this thesis.


\subsection{Tikhonov regularization}
% Introduce general min problem, and normal equation.
% Then introduce Tikhonov 
% Then introduce general Tikhonov
Let $K$ be an $N_b \times N$ matrix, and consider the problem
\begin{equation}\label{eq:tikhonov}
    \min_{x \in \mathbb R^N} \| K x - y \|^2 + \lambda^2 \| W x \|^2,
\end{equation}
where $\lambda \ge 0$ is the regularization parameter and $W \in \mathbb{R}^{N \times N}$ are the regularization weights. The solution $x_\lambda$ of Eq.~\eqref{eq:tikhonov} is known as the Tikhonov solution of $K x = y$. The term $\| W x \|^2$ punishes the weighted norm $x$ preventing the norm of $x$ from exploding, which often is a problem in ill-posed problems \textcolor{red}{(reference)}.


\subsection{Randomized SVD}
The rSVD is a technique used to compute a rank-$k$ approximation of an $m\times n$ matrix without computing the SVD. The core idea is to use random sampling to find a low-rank matrix $Q$ which approximates the range of $A$, project $A$ onto $\operatorname{span}(Q)$, and compute the SVD of the projection. The projection $B=Q^TA$ will, with high probability, capture the dominant actions of $A$, and the SVD of $B$ will yield a good approximation \cite{halko_finding_2011}. This procedure is summarized in Algorithm \ref{alg:rsvd}. The construction of an orthonormal basis $Q$ of $Y$ is done through a QR factorization. Note that the approximation $A\approx QQ^T A$ and the approximate factorization $A\approx U\Sigma V^T$ produced by rSVD have the same error. This can be seen by observing that $QQ^TA = QB = Q\tilde U\Sigma V^T$, and setting $U=Q\tilde U$.
\begin{algorithm}[!htbp]
\caption{rSVD}
\label{alg:rsvd}
\begin{algorithmic}[1]
    \Require $A\in\mathbb R^{m\times n}$, target rank $k$, oversampling parameter $p$.
    \Ensure Approximate rank-$k$ factorization $U\Sigma V^T$.
    \State Generate a random $n\times(k+p)$ test matrix $\Omega$.
    \State Compute $Y = A\Omega$.
    \State Form an orthonormal basis $Q$ of $Y$.
    \State Set $B = Q^TA$.
    \State Compute the SVD $B = \tilde U\Sigma V^T$.
    \State Set $U = Q\tilde U$. 
\end{algorithmic}
\end{algorithm}


\subsection{Tips}
When you are closely following a book to explain something, which is often the case in a theory section, you can write at the start of the section you are about to introduce: ``This section follows closely the reference book by Sumiyoshi Abe and Yuko Okamoto, \emph{Nonextensive Statistical Mechanics and Its Applications}~\parencite[p.~32]{Abe2001nonextensive}.''

A Master's~\cite{LastName2045norwegian} or a PhD thesis~\cite{Temult2038binding} should include the name of the university wherein it was written, as well as the year. Moreover, it should include a URL to the work, when available.

\clearpage
\section{Methods}
\subsection{Matrix-free randomized SVD}
\input{figures/rsvd.tex}
Let $K$ be the $N_b \times N$ discrete forward operator given in Eq.~\eqref{eq:K}. We are interested in a rank-$k$ approximation $K_k$ of the matrix $K$. By the Eckart–Young–Mirsky theorem, the optimal rank-$k$ approximation of $K$ is the truncated SVD obtained by keeping the $k$ first singular values, and their corresponding singular values. The randomized SVD provides an efficient alternative, by computing an approximate rank-$k$ factorization $K \approx U_k \Sigma_k V_k$, which also approximates the optimal truncated rank-$k$ SVD decomposition. We will now present a matrix-free variation of the rSVD algorithm, suited for problems where the matrix $K$ is not explicitly known.
\nomenclature[G]{$k$}{Target rank}

Like the rSVD, this matrix-free formulation consists of two stages. In the first stage, an approximate basis $Q$ of $K$ is constructed through random sampling of the column space of $K$. In the second stage, the matrix $K$ is projected onto the subspace $\operatorname{span}(Q)$, and a SVD factorization is computed of the projected matrix.

In the first stage, we apply $K$ on $k$ random vectors $\psi_1, \psi_2, \dots, \psi_l$ drawn from some distribution, typically the normal distribution. This yields the random samples $y_1, y_2, \dots, y_k$, where $y_i = K \psi_i$. If we gather the samples and random vectors into matrices $Y = [y_i]_{i=1}^k$ and $\Psi = [\psi_i]_{i}$, we can express $Y$ as
\begin{equation}
    Y = K \Psi.
\end{equation}
(something about how $y_i$ is most likely to line up with the first singular vector). Geometrically, we want the $i$'th sample $y_i$ to approximately point in the direction of the the $i$ right singular vector $v_i$. To remove contributions of the previous right singular vectors $v_{i-1}, v_{i-2}, \dots, v_{1}$ on $y_i$, we must orthogonalize the random samples. A QR-factorization
\begin{equation}
    QR = Y
\end{equation}
achieves such an orthogonalization, obtaining the matrix $Q$ which contains the orthogonalized column-vectors $q_1, q_2, \dots, q_k$. This concludes the first step, where we have obtained an approximation of the column space of $K$:
\begin{equation}
    \operatorname{span}(Q) \approx \operatorname{span}(K)
\end{equation}
\nomenclature[G]{$Y$}{Sample matrix}
\nomenclature[G]{$\Psi$}{Random matrix}

For the second stage, we project $K$ down to the subspace $\operatorname{span}(Q)$. This is typically done by computing $B = Q^T K$. However, since we do not know $K$ explicitly, we instead compute $B^* = K^* Q$, and then $B$ can be found by computing $B = (B^*)^*$.

To do this in practice, we use the the definition of $K^*$ under the finite element space $V_h$,
\begin{equation}\label{eq:K_star}
    (K^* z)^T M_{dx} = z^T M_{ds} K.
\end{equation}
If we choose $z = M_{ds}^{-1} q_i$ and insert it into Eq.~\eqref{eq:K_star}, we see that
\begin{equation}\label{eq:rsvd_relation}
    (K^* z)^T M_{dx}
    = (M_{ds}^{-1} q_i)^T M_{ds} K
    = q_i^T K.
\end{equation}
Hence, if we let
\begin{equation}\label{eq:rsvd_adjoint_problem}
    b_i = (K^* (M_{ds}^{-1} q_i))^T M_{dx}
\end{equation}
for $i = 1, 2, \dots, k$, and define the $k \times N$ matrix
\begin{equation}
    B = \begin{bmatrix} b_1 \\ \vdots \\ b_k \end{bmatrix}
\end{equation}
it must be that, by Eq.~\eqref{eq:rsvd_relation},
\begin{equation}
    B = Q^T K,
\end{equation}
as desired.

We have computed the projection $B = Q^T K$ by solving the adjoint problem in Eq.~\eqref{eq:rsvd_adjoint_problem} $k$ times. In the final part of the second stage, we take the SVD of $B$,
\begin{equation}
    B = \tilde U_k \Sigma_k V_k^T,
\end{equation}
and set $U_k = Q \tilde U_k$. This gives an approximate SVD decomposition of the form
\begin{equation}\label{eq:rsvd_approx}
    K \approx U_k \Sigma_k V_k^T.
\end{equation}
By noting that $U_k \Sigma_k V_k^T = Q \tilde U_k \Sigma_k V_k^T = Q B = Q Q^T K$, and that $\operatorname{rank}(Q) = k$, we know that the factorization in Eq.~\eqref{eq:rsvd_approx} is a rank-$k$ approximation of $K$. The procedure is summarized in Algorithm~\ref{alg:discrete_rsvd}.


\begin{algorithm}[!thbp]
\caption{Discrete Operator rSVD Approximation of $K$}
\label{alg:discrete_rsvd}
\begin{algorithmic}[1]
    \Require Forward operator $K: \mathbb{R}^N \to \mathbb{R}^{N_b}$, adjoint operator $K^*: \mathbb{R}^{N_b} \to \mathbb{R}^N$, mass matrices $M_{dx}, M_{ds}$, target rank $k$
    \Ensure Approximate rank-$k$ factorization $K \approx U \Sigma V^T$
    
    \State \textbf{Range sketch:}
    \For{$i = 1, \dots, k$}
        \State Draw a random vector $\psi_i \in \mathbb{R}^N$
        \State Apply the forward operator: $y_i = K \psi_i \in \mathbb{R}^{N_b}$
    \EndFor

    \State Form the matrix $Y = [y_1, \dots, y_k] \in \mathbb{R}^{N_b \times k}$ \Comment{Equivalent to $Y = K \Psi$}
    \State Compute an orthonormal basis: $Q = \text{orth}(Y)$

    \State \textbf{Projection onto the basis:}
    \For{$i = 1, \dots, p$}
        \State Let $q_i$ be the $i$-th column of $Q$
        \State Apply the adjoint operator: $\tilde b_i = K^* (M_{ds}^{-1} q_i)$
        \State Set $b_i = \tilde b_i^T M_{dx}$ \Comment{$i$-th row of $B$}
    \EndFor
    \State Form the projected matrix $B = \begin{bmatrix} b_1 \\ \vdots \\ b_k \end{bmatrix} \in \mathbb{R}^{k \times N}$ \Comment{Equivalent to $B = Q^T Y$}

    \State Compute the SVD: $B = \tilde U \Sigma V^T$
    \State Set $U = Q \tilde U$
    
\end{algorithmic}
\end{algorithm}


\clearpage
\subsection{Weighted Tikhonov using rSVD}
Let $K_k = U_k \Sigma_k V_k^T$ be the rank-$k$ approximate SVD of $K$ obtained using Algorithm~\ref{alg:discrete_rsvd}. The weights $W = \operatorname{diag}(w_i)_{i=1}^N$ \textcolor{red}{(referer tilbake til metoden Elvetun \cite{elvetun_regularization_2020})}, i.e.,
\begin{equation}
    w_i = \frac
    { \sqrt{y_i^T M_{dx} y_i} }
    { \int_{\Omega} \phi_i \, dx }, \quad
    y_i = K^+ K e_i,
\end{equation}
where $K^+$ denotes the pseudoinverse of $K$. We may now replace $K$ with its rank-$k$ approximate factorization $U_k \Sigma_k V_k^T$ to obtain
\begin{equation}
    y_i = V_k V_k^T e_i.
\end{equation}



\clearpage
\subsection{Dynamic low-rank solver}
Let $X = \operatorname{matrix}(x)$ and $X = U_x \Sigma_x V_x^T$ be the SVD of $X$ and let $\Phi(x)$ be some differentiable cost function of $x$, and $D = \operatorname{matrix}(\nabla \Phi(x))$ its matrix equivalent. We wish to update $X^{(k)}$ according to
$$
X^{(k+1)} = X^{(k)}-\alpha D
$$
Inserting the SVD of $X^{(k)}$ at step $k$, $X^{(k)} = U_x^{(k)} \Sigma_x^{(k)} (V_x^{(k)})^T$, we get that
$$
U_x^{(k+1)} \Sigma_x^{(k+1)} (V_x^{(k+1)})^T = U_x^{(k)} \Sigma_x^{(k)} (V_x^{(k)})^T - \alpha D. \quad (1)
$$
If we right-multiply by $V_x^{(k)}$ and let $W^{(k)} = U_x^{(k)} \Sigma_x^{(k)}$, we get
$$
W^{(k+1)}(V_x^{(k+1)})^T(V_x^{(k)}) = W^{(k)} - \alpha D V_x^{(k)}.
$$
The approximation $(V_x^{(k+1)})^T(V_x^{(k)}) \approx I$ suggests the following updating scheme:
$$
W^{(k+1)} = W^{(k)} - \alpha D V_x^{(k)}.
$$
If we transpose  Equation $(1)$, right-multiply by $U_x^{(k)}$ and denote $L^{(k)} = V_x^{(k)} (\Sigma_x^{(k)})^T$, we get a similar updating scheme for $L^{(k)}$,
$$
L^{(k+1)} = L^{(k)} - \alpha D^T U_x^{(k)}.
$$
We now have update schemes for $W^{(k)} = U_x^{(k)} \Sigma_x^{(k)}$ and $L^{(k)} = V_x^{(k)} (\Sigma_x^{(k)})^T$. Further, we expand and orthogonalize the basis' $U^{(k)}$ and $V^{(k)}$ with the updated weighted basis' $W^{(k+1)}$ and $L^{(k+1)}$, respectively, in the following way:
$$
\begin{aligned}
\hat U &= \operatorname{orth}\begin{bmatrix} U_x^{(k)} & W^{(k+1)}\end{bmatrix}, \\
\hat V &= \operatorname{orth}\begin{bmatrix} V_x^{(k)} & L^{(k+1)}\end{bmatrix}.
\end{aligned}
$$
Currently, we have obtained an approximation $X \approx \hat U \Sigma_x \hat V^T$. In terms of $\Sigma_x$, we have
$$
\Sigma_x \approx \hat U^T X \hat V.
$$
Substituting $X = U_x \Sigma_x V_x^T$ gives the approximation $\hat \Sigma^{(k)}$ of $\Sigma_x$,
$$
\Sigma_x \approx \hat \Sigma^{(k)} =(\hat U^T U_x)\Sigma_x(V_x^T \hat V). \quad (2)
$$
Now, if we write the gradient descent scheme $X^{(k+1)} = X^{(k)} - \alpha D$ as
$$
\hat U^T X^{(k+1)} \hat V = \hat U^TX^{(k)}\hat V - \alpha \hat U^T D \hat V,
$$
and use Equation $(2)$, we get the following updating scheme for $\hat \Sigma^{(k)}$,
$$
\hat \Sigma^{(k+1)} = \hat \Sigma^{(k)} - \alpha \hat U^T D \hat V.
$$
The final step is then to truncate these updated matrices to obtain
$$
U_x^{(k+1)}, \Sigma_x^{(k+1)}, V_x^{(k+1)} = \operatorname{truncate}(\hat U, \hat \Sigma, \hat V)
$$
and finally, set $X^{(k+1)} = U_x^{(k+1)} \Sigma_x^{(k+1)} (V_x^{(k+1)})^T$.


\clearpage
\section{Literature Review}


\clearpage
\section{Methodology}
In the methodology, you might include mentions of various \texttt{python} (or other programming languages) packages.


\clearpage
\section{Results}
\subsection{Forward operator computational time}
\input{figures/exact_vs_rsvd_comp_time.tex}

\subsection{Weights computational time}

\clearpage
\section{Discussion}

\clearpage
\section{Conclusion}

% Note that the appendices should come after the bibliography, not before.
\clearpage
\references % This command prints out your bibliography

\clearpage
\appendices % This command is needed to start the appendix section and number the figures and tables accordingly. You should still use "\section"s to make your appendices.
\renewcommand{\theequation}{A\arabic{equation}}
\setcounter{equation}{0}

\section{Mathematical preliminaries}
\subsection{Function spaces}\label{appendix:function_spaces}
Let $\Omega \subset \mathbb{R}^d$ be an open set and $\mathbb{F}$ denote either $\mathbb{R}$ or $\mathbb{C}$. The space $L^2(\Omega)$, often referred to as the Lebesgue space of square integrable functions, is defined as 
\begin{equation}
    L^2(\Omega) =
    \left\{
        f : \Omega \to \mathbb{F} \, \Bigg| \,
        \int_{\Omega} |f(x)|^2 \, dx < \infty
    \right\},
\end{equation}
with inner product
\begin{equation}\label{eq:L2_inner_product}
    \langle f,\, g \rangle_{L^2(\Omega)} = \int_\Omega f(x) \overline{g(x)} \, dx.
\end{equation}
The space $L^2(\Omega)$ is a Hilbert space, i.e., a complete inner product space. For details, see Section~3 in Kreyszig \cite{kreyszigIntroductoryFunctionalAnalysis1989}.

The space $H^1(\Omega)$, known as the first-order Sobolev space, is defined as
\begin{equation}
    H^1(\Omega) =
    \left\{
        f \in L^2(\Omega) \, \Bigg| \,
        \frac{\partial f}{\partial x_i} \in L^2(\Omega),\, i = 1, \dots, d  
    \right\},
\end{equation}
with inner product
\begin{equation}
    \langle f,\, g \rangle_{H^1(\Omega)} = 
    \langle f,\, g \rangle_{L^2(\Omega)} +
    \langle \nabla f,\, \nabla g \rangle_{L^2(\Omega)}.
\end{equation}
The space $H^1(\Omega)$ is a Hilbert space and a subspace of $L^2(\Omega)$. See Section~2 in Quarteroni for further details \cite{quarteroniNumericalModelsDifferential2014}.


\subsection{Weak derivative}
Let $\Omega \subset \mathbb{R}^d$ be an open set and $u \in L^2(\Omega)$. A function $v$ is said to be the weak derivative of $u$ if
\begin{equation}
    \int_\Omega u(x) \frac{\partial \phi}{\partial x_i} \, dx =
    - \int_\Omega v \phi \, dx
\end{equation}
for all infinitely differentiable $\phi$ with compact support on $\Omega$. For details, see Section~2 in Quarteroni \cite{quarteroniNumericalModelsDifferential2014}.


\subsection{The stiffness and the mass matrix}
Let $V_h \subset H^1(\Omega)$ be the finite element space defined in Eq.~\eqref{eq:fem_space}. Under the basis $\{\phi_i\}_{i=1}^N$, the stiffness matrix $A \in \mathbb{R}^{N \times N}$ and mass matrix $M \in \mathbb{R}^{N \times N}$ are defined as
\begin{equation}
    A_{ij} = \int_\Omega 
    \left( \sigma \nabla \phi_j \cdot \nabla \phi_i 
    + k \phi_j \phi_i \right) dx,
    \quad
    M_{ij} = \int_\Omega \phi_j \phi_i \, dx.
\end{equation}
Both matrices are symmetric. Under the standing assumptions that $\sigma$ is positive definite and $c \ge 0$, the stiffness matrix $A$ is positive definite and therefore invertible. 

In addition, both $A$ and $M$ are sparse matrices, since $A_{ij}$ and $M_{ij}$ are non-zero only if $\phi_i$ and $phi_j$ share a element in the mesh.

\subsection{Positive definiteness}\label{sec:positive_definiteness}
Let $\sigma = \sigma(x)$ be a symmetric $N \times N$ matrix. We say that $\sigma(x)$ is positive definite if
\begin{equation}
    y^T \sigma(x) y \ge 0
\end{equation}
for all $y \in \mathbb{R}^N$.

\end{document}
